---
title: "miR-100 analysis"
author: "Jill Ashey"
date: "2026-06-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

Read in miranda data 
```{r}
## Apoc 
cds_apoc <- read.delim("../output/miRNA_interactions/apoc/miranda_strict_mRNA_parsed_apoc.txt", header = F)[,c(1,2,7)]
colnames(cds_apoc) <- c("miRNA", "mRNA", "total_bp_shared")
cds_apoc$Region <- "CDS"
utr_apoc <- read.csv("../output/miRNA_interactions/apoc/miranda_strict_3UTR_parsed_genename_apoc.csv", header = T)[,c(2,8,11)]
colnames(utr_apoc) <- c("miRNA", "total_bp_shared", "mRNA")
utr_apoc$Region <- "3UTR"
binding_apoc <- rbind(cds_apoc, utr_apoc)
binding_apoc$miRNA <- sub("^>", "", binding_apoc$miRNA) 

## Ahya 
cds_ahya <- read.delim("../output/miRNA_interactions/ahya/miranda_strict_mRNA_parsed_ahya.txt", header = F)[,c(1,2,7)]
colnames(cds_ahya) <- c("miRNA", "mRNA", "total_bp_shared")
cds_ahya$Region <- "CDS"
utr_ahya <- read.csv("../output/miRNA_interactions/ahya/miranda_strict_3UTR_parsed_genename_ahya.csv", header = T)[,c(2,8,11)]
colnames(utr_ahya) <- c("miRNA", "total_bp_shared", "mRNA")
utr_ahya$Region <- "3UTR"
binding_ahya <- rbind(cds_ahya, utr_ahya)
binding_ahya$miRNA <- sub("^>", "", binding_ahya$miRNA) 

## Nvec 
cds_nvec <- read.delim("../output/miRNA_interactions/nvec/miranda_strict_mRNA_parsed_nvec.txt", header = F)[,c(1,2,7)]
colnames(cds_nvec) <- c("miRNA", "mRNA", "total_bp_shared")
cds_nvec$Region <- "CDS"
utr_nvec <- read.csv("../output/miRNA_interactions/nvec/miranda_strict_3UTR_parsed_genename_nvec.csv", header = T)[,c(2,8,11)]
colnames(utr_nvec) <- c("miRNA", "total_bp_shared", "mRNA")
utr_nvec$Region <- "3UTR"
binding_nvec <- rbind(cds_nvec, utr_nvec)
binding_nvec$miRNA <- sub("^>", "", binding_nvec$miRNA) 
```

Subset miR-100 from each dataset
```{r}
# Apoc - Cluster_1200 is miR-100
mir100_binding_apoc <- binding_apoc %>%
  filter(miRNA == "Cluster_1200")

# Ahya - Cluster_1356 is miR-100
mir100_binding_ahya <- binding_ahya %>%
  filter(miRNA == "Cluster_1356")

# Nvec - Cluster_1430 is miR-100
mir100_binding_nvec <- binding_nvec %>%
  filter(miRNA == "Cluster_1430")
```

How many unique mRNAs does miR-100 bind?
```{r}
length(unique(mir100_binding_apoc$mRNA))
length(unique(mir100_binding_ahya$mRNA))
length(unique(mir100_binding_nvec$mRNA))
```

Count how many targets are in the CDS, 3UTR or both
```{r}
# Apoc
cds_genes_apoc  <- unique(mir100_binding_apoc$mRNA[mir100_binding_apoc$Region == "CDS"])
utr_genes_apoc  <- unique(mir100_binding_apoc$mRNA[mir100_binding_apoc$Region == "3UTR"])
both_genes_apoc <- intersect(cds_genes_apoc, utr_genes_apoc)
cat("Apoc: Exclusive to CDS:", length(setdiff(cds_genes_apoc, utr_genes_apoc)), "\n",
    "Apoc: Exclusive to 3UTR:", length(setdiff(utr_genes_apoc, cds_genes_apoc)), "\n",
    "Apoc: Targeted in BOTH:", length(both_genes_apoc), "\n")

# Ahya
cds_genes_ahya  <- unique(mir100_binding_ahya$mRNA[mir100_binding_ahya$Region == "CDS"])
utr_genes_ahya  <- unique(mir100_binding_ahya$mRNA[mir100_binding_ahya$Region == "3UTR"])
both_genes_ahya <- intersect(cds_genes_ahya, utr_genes_ahya)
cat("Ahya: Exclusive to CDS:", length(setdiff(cds_genes_ahya, utr_genes_ahya)), "\n",
    "Ahya: Exclusive to 3UTR:", length(setdiff(utr_genes_ahya, cds_genes_ahya)), "\n",
    "Ahya: Targeted in BOTH:", length(both_genes_ahya), "\n")

# Nvec
cds_genes_nvec  <- unique(mir100_binding_nvec$mRNA[mir100_binding_nvec$Region == "CDS"])
utr_genes_nvec  <- unique(mir100_binding_nvec$mRNA[mir100_binding_nvec$Region == "3UTR"])
both_genes_nvec <- intersect(cds_genes_nvec, utr_genes_nvec)
cat("Nvec: Exclusive to CDS:", length(setdiff(cds_genes_nvec, utr_genes_nvec)), "\n",
    "Nvec: Exclusive to 3UTR:", length(setdiff(utr_genes_nvec, cds_genes_nvec)), "\n",
    "Nvec: Targeted in BOTH:", length(both_genes_nvec), "\n")
```

Plot binding percentages across mRNA, 3' UTR or both. 
```{r}
# Data frame with counts
percent <- data.frame(
  Species = rep(c("Acropora hyacinthus", "Astrangia poculata", "Nematostella vectensis"), each = 3),
  Binding = rep(c("CDS only", "3' UTR only", "Both"), times = 3),
  Count = c(242, 140, 0,
            319, 136, 1,
            900, 103, 0))

# Calculate percentages within each species
percent <- percent %>%
  group_by(Species) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

# Calculate total counts per species for labels
totals <- percent %>%
  group_by(Species) %>%
  summarize(Total = sum(Count))

# Plot stacked bar plot with total counts on top of each bar
feature_plot <- ggplot(percent, aes(x = Species, y = Percent, fill = Binding)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", position = position_stack(reverse = F)) +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  # geom_text(
  #   data = totals,
  #   aes(x = Species, y = 105, label = Total),
  #   inherit.aes = FALSE,   
  #   size = 4) +
  scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 100, by = 10),
        limits = c(0, 105)) +
  ylab("Percentage (%)") +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  coord_cartesian(clip = "off") +
  ylab("Percentage (%)") +
  xlab("") +
  theme_classic() + 
  theme(axis.text = element_text(color = "black", size = 21),
    axis.title = element_text(size = 23, face = "bold"),
    axis.text.x = element_text(face = "italic"),
    legend.title = element_text(size = 20, face = "bold"),
    legend.text = element_text(size = 19),
    legend.key.size = unit(0.6, "cm")) +
  guides(fill = guide_legend(reverse = F)); feature_plot

ggsave("../output/miRNA_interactions/shared/mir100-target_binding_feature.pdf", plot = feature_plot, width = 10, height = 8, dpi = 300)
ggsave("../output/miRNA_interactions/shared/mir100-target_binding_feature.png", plot = feature_plot, width = 10, height = 8, dpi = 300)
```

## Target homology of miR-100

Read in broccoli results 
```{r}
broc_groups <- data.table::fread("../output/broccoli/dir_step3/table_OGs_protein_names.txt", header = T)
```

Identify targets in Broccoli data 
```{r}
# Function to extract gene IDs from protein IDs
extract_gene_id <- function(protein_ids, species) {
  if (species == "nvec") {
    # Remove "protein|" prefix if present, keep the base ID
    gsub("protein\\|", "", protein_ids)
  } else if (species == "ahya") {
    # Keep as is
    protein_ids
  } else if (species == "apoc") {
    # Remove "protein|" prefix if present
    gsub("protein\\|", "", protein_ids)
  }
}

# Extract gene IDs from the broc_groups columns
broc_groups[, Apoc_genes := lapply(strsplit(Apoc_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "apoc"))]
broc_groups[, Nvec_genes := lapply(strsplit(Nvec_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "nvec"))]
broc_groups[, Ahya_genes := lapply(strsplit(Ahya_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "ahya"))]
```

Subset genes that are targeted by miR-100
```{r}
targets_ahya <- unique(mir100_binding_ahya$mRNA)
targets_apoc <- unique(mir100_binding_apoc$mRNA)
targets_nvec <- unique(mir100_binding_nvec$mRNA)

filtered_broc <- broc_groups[
  mapply(function(apoc, nvec, ahya) {
    any(apoc %in% targets_apoc) | 
    any(nvec %in% targets_nvec) | 
    any(ahya %in% targets_ahya)
  }, Apoc_genes, Nvec_genes, Ahya_genes)
]
```

Add presence/absence cols 
```{r}
# Drop the protein cols 
cols_to_remove <- c("Apoc_proteins.faa", "Nvec_proteins.faa", "Ahya_proteins.faa")
filtered_broc[, (cols_to_remove) := NULL]

filtered_broc[, present_apoc := sapply(Apoc_genes, function(x) length(x) > 0 && !all(x == "" | is.na(x)))]
filtered_broc[, present_nvec := sapply(Nvec_genes, function(x) length(x) > 0 && !all(x == "" | is.na(x)))]
filtered_broc[, present_ahya := sapply(Ahya_genes, function(x) length(x) > 0 && !all(x == "" | is.na(x)))]
```

Filter for rows where all spp have a miR-100 target in that orthogroup 
```{r}
filtered_broc <- filtered_broc[present_apoc == TRUE & 
                                present_nvec == TRUE & 
                                present_ahya == TRUE]
```

## Apoc miR-100 target enrichment 

### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
apoc_gene2go<-read.delim("../data/annotations/Apoc_gene2go.tab", sep="\t")
```

Make list of genes that are present in the orthogroups in all 3 spp AND are a miR-100 target
```{r}
# Unlist the Apoc genes from these specific conserved orthogroups
apoc_genes_for_enrichment <- unique(unlist(filtered_broc$Apoc_genes))

# CRITICAL: Keep only the ones that are actually miR-100 targets
apoc_final_query <- apoc_genes_for_enrichment[apoc_genes_for_enrichment %in% targets_apoc]
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
apoc_clust_genes <- as.character(unique(apoc_final_query))

# All genes 
apoc_all_genes <- as.character(apoc_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
apoc_GeneList <- factor(as.integer(apoc_all_genes %in% apoc_clust_genes))
names(apoc_GeneList) <- apoc_all_genes
```

Read in gene-to-go-mappings
```{r}
apoc_gene2go_topgo<-readMappings("../data/annotations/Apoc_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APOC_BP <-new("topGOdata", ontology="BP", gene2GO=apoc_gene2go_topgo, allGenes=apoc_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APOC_BP_FE <- runTest(GO_APOC_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APOC_BP_En <- GenTable(GO_APOC_BP, Fisher = GO_APOC_BP_FE, orderBy = "Fisher", numChar = 51, topNodes = 200)
```

Filter by significant results
```{r}
GO_APOC_BP_En$Fisher<-as.numeric(GO_APOC_BP_En$Fisher)
GO_APOC_BP_En_sig<-GO_APOC_BP_En[GO_APOC_BP_En$Fisher<0.05,]
```

### Molecular Functions 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APOC_MF <-new("topGOdata", ontology="MF", gene2GO=apoc_gene2go_topgo, allGenes=apoc_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APOC_MF_FE <- runTest(GO_APOC_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APOC_MF_En <- GenTable(GO_APOC_MF, Fisher = GO_APOC_MF_FE, orderBy = "Fisher", numChar = 51, topNodes = 200)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APOC_MF_En$Fisher<-as.numeric(GO_APOC_MF_En$Fisher)
GO_APOC_MF_En_sig<-GO_APOC_MF_En[GO_APOC_MF_En$Fisher<0.05,]
```

### Plotting BP + MF terms for Apoc miR-100 binding 

Plot top 10 terms
```{r}
## Bind dfs 
GO_APOC_BP_En_sig$ontology <- "Biological Processes"
GO_APOC_MF_En_sig$ontology <- "Molecular Functions"
BP_MF_APOC <- rbind(GO_APOC_BP_En_sig, GO_APOC_MF_En_sig)

## Select top 10 terms for each ontology 
top10_per_ontology_apoc <- BP_MF_APOC %>%
  group_by(ontology) %>%
  slice_max(Significant, n = 10, with_ties = FALSE) %>%
  ungroup()
  
ggplot(top10_per_ontology_apoc, aes(x = reorder(Term, Significant), y = Significant, fill = Fisher)) +
  geom_bar(stat = "identity") +
  #geom_text(aes(label = gene_count), vjust = -0.5, color = "black", size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  labs(x = "GO Term",
       y = "Number of Significant Genes",
       fill = "Fisher Value") +
  theme_bw()
```

## Ahya miR-100 target enrichment 

### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
ahya_gene2go<-read.delim("../data/annotations/Ahya_gene2go.tab", sep="\t")
```

Make list of genes that are present in the orthogroups in all 3 spp AND are a miR-100 target
```{r}
# Unlist the Apoc genes from these specific conserved orthogroups
ahya_genes_for_enrichment <- unique(unlist(filtered_broc$Ahya_genes))

# CRITICAL: Keep only the ones that are actually miR-100 targets
ahya_final_query <- ahya_genes_for_enrichment[ahya_genes_for_enrichment %in% targets_ahya]
ahya_final_query
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
ahya_clust_genes <- as.character(unique(ahya_final_query))

# All genes 
ahya_all_genes <- as.character(ahya_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
ahya_GeneList <- factor(as.integer(ahya_all_genes %in% ahya_clust_genes))
names(ahya_GeneList) <- ahya_all_genes
```

Read in gene-to-go-mappings
```{r}
ahya_gene2go_topgo<-readMappings("../data/annotations/Ahya_gene2go.tab", IDsep=";", sep="\t")
```

### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_AHYA_BP <-new("topGOdata", ontology="BP", gene2GO=ahya_gene2go_topgo, allGenes=ahya_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_AHYA_BP_FE <- runTest(GO_AHYA_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_AHYA_BP_En <- GenTable(GO_AHYA_BP, Fisher = GO_AHYA_BP_FE, orderBy = "Fisher", numChar = 51, topNodes = 200)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_AHYA_BP_En$Fisher<-as.numeric(GO_AHYA_BP_En$Fisher)
GO_AHYA_BP_En_sig<-GO_AHYA_BP_En[GO_AHYA_BP_En$Fisher<0.05,]
```

### Molecular Functions 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_AHYA_MF <-new("topGOdata", ontology="MF", gene2GO=ahya_gene2go_topgo, allGenes=ahya_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_AHYA_MF_FE <- runTest(GO_AHYA_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_AHYA_MF_En <- GenTable(GO_AHYA_MF, Fisher = GO_AHYA_MF_FE, orderBy = "Fisher", numChar = 51, topNodes = 200)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_AHYA_MF_En$Fisher<-as.numeric(GO_AHYA_MF_En$Fisher)
GO_AHYA_MF_En_sig<-GO_AHYA_MF_En[GO_AHYA_MF_En$Fisher<0.05,]
```

### Plotting BP + MF terms for Ahya miR-100 binding 

Plot top 10 terms
```{r}
## Bind dfs 
GO_AHYA_BP_En_sig$ontology <- "Biological Processes"
GO_AHYA_MF_En_sig$ontology <- "Molecular Functions"
BP_MF_AHYA <- rbind(GO_AHYA_BP_En_sig, GO_AHYA_MF_En_sig)

## Select top 10 terms for each ontology 
top10_per_ontology_ahya <- BP_MF_AHYA %>%
  group_by(ontology) %>%
  slice_max(Significant, n = 10, with_ties = FALSE) %>%
  ungroup()
  
ggplot(top10_per_ontology_ahya, aes(x = reorder(Term, Significant), y = Significant, fill = Fisher)) +
  geom_bar(stat = "identity") +
  #geom_text(aes(label = gene_count), vjust = -0.5, color = "black", size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  labs(x = "GO Term",
       y = "Number of Significant Genes",
       fill = "Fisher Value") +
  theme_bw()
```

## Nvec miR-100 target enrichment 

### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
nvec_gene2go<-read.delim("../data/annotations/Nvec_gene2go.tab", sep="\t")
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
nvec_clust_genes <- as.character(unique(mir100_binding_nvec$mRNA))

# All genes 
nvec_all_genes <- as.character(nvec_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
nvec_GeneList <- factor(as.integer(nvec_all_genes %in% nvec_clust_genes))
names(nvec_GeneList) <- nvec_all_genes
```

Read in gene-to-go-mappings
```{r}
nvec_gene2go_topgo<-readMappings("../data/annotations/Nvec_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_NVEC_BP <-new("topGOdata", ontology="BP", gene2GO=nvec_gene2go_topgo, allGenes=nvec_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_NVEC_BP_FE <- runTest(GO_NVEC_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_NVEC_BP_En <- GenTable(GO_NVEC_BP, Fisher = GO_NVEC_BP_FE, orderBy = "Fisher", numChar = 51, topNodes = 200)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_NVEC_BP_En$Fisher<-as.numeric(GO_NVEC_BP_En$Fisher)
GO_NVEC_BP_En_sig<-GO_NVEC_BP_En[GO_NVEC_BP_En$Fisher<0.05,]
```

### Molecular Functions 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_NVEC_MF <-new("topGOdata", ontology="MF", gene2GO=nvec_gene2go_topgo, allGenes=nvec_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_NVEC_MF_FE <- runTest(GO_NVEC_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_NVEC_MF_En <- GenTable(GO_NVEC_MF, Fisher = GO_NVEC_MF_FE, orderBy = "Fisher", numChar = 51, topNodes = 200)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_NVEC_MF_En$Fisher<-as.numeric(GO_NVEC_MF_En$Fisher)
GO_NVEC_MF_En_sig<-GO_NVEC_MF_En[GO_NVEC_MF_En$Fisher<0.05,]
```

### Plotting BP + MF terms for Nvec miR-100 binding 

Plot top 10 terms
```{r}
## Bind dfs 
GO_NVEC_BP_En_sig$ontology <- "Biological Processes"
GO_NVEC_MF_En_sig$ontology <- "Molecular Functions"
BP_MF_NVEC <- rbind(GO_NVEC_BP_En_sig, GO_NVEC_MF_En_sig)

## Select top 10 terms for each ontology 
top10_per_ontology_nvec <- BP_MF_NVEC %>%
  group_by(ontology) %>%
  slice_max(Significant, n = 10, with_ties = FALSE) %>%
  ungroup()
  
ggplot(top10_per_ontology_nvec, aes(x = reorder(Term, Significant), y = Significant, fill = Fisher)) +
  geom_bar(stat = "identity") +
  #geom_text(aes(label = gene_count), vjust = -0.5, color = "black", size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  labs(x = "GO Term",
       y = "Number of Significant Genes",
       fill = "Fisher Value") +
  theme_bw()
```

## Enriched terms for miR-100 targets in all app

Find shared enriched terms 
```{r}
# Add species labels and keep only needed columns
ahya <- BP_MF_AHYA %>%
  mutate(species = "Acropora hyacinthus") %>%
  dplyr::select(species, GO.ID, Term, Fisher, Significant, ontology)

apoc <- BP_MF_APOC %>%
  mutate(species = "Astrangia poculata") %>%
  dplyr::select(species, GO.ID, Term, Fisher, Significant, ontology)

nvec <- BP_MF_NVEC %>%
  mutate(species = "Nematostella vectensis") %>%
  dplyr::select(species, GO.ID, Term, Fisher, Significant, ontology)

# Shared GO IDs across all three
shared_ids <- purrr::reduce(
  list(ahya$GO.ID, apoc$GO.ID, nvec$GO.ID),
  intersect)
shared_ids
```

No shared GO terms for miR-100! Interesting 

Plot individual go terms for each spp 
```{r}
all_spp_go <- rbind(apoc, ahya, nvec)

# Filter for top 10 for each spp 
top_go <- all_spp_go %>%
  group_by(species) %>%
  slice_min(order_by = Fisher, n = 10, with_ties = FALSE) %>%
  ungroup()

# Plot A: Ahya
p1 <- top_go %>% 
  filter(species == "Acropora hyacinthus") %>% 
  ggplot(aes(x = -log10(Fisher), y = reorder(Term, -Fisher))) +
  geom_point(aes(shape = ontology), size = 4, color = "#E69F00") +
  theme_bw() +
  facet_wrap(~species) +
  labs(x = NULL, y = "GO Term", shape = "Ontology") + # x = NULL hides label for upper panels
  theme(plot.title = element_text(face = "bold")); p1

# Plot B: Apoc
p2 <- top_go %>% 
  filter(species == "Ahya") %>% 
  ggplot(aes(x = -log10(Fisher), y = reorder(Term, -Fisher))) +
  geom_point(aes(shape = ontology), size = 4, color = "#56B4E9") +
  theme_bw() +
  labs(title = "Ahya", x = NULL, y = "GO Term", shape = "Ontology") +
  theme(plot.title = element_text(face = "bold"))

# Plot C: Nvec
p3 <- top_go %>% 
  filter(species == "Nvec") %>% 
  ggplot(aes(x = -log10(Fisher), y = reorder(Term, -Fisher))) +
  geom_point(aes(shape = ontology), size = 4, color = "#009E73") +
  theme_bw() +
  labs(title = "Nvec", x = "-log10(Fisher p-value)", y = "GO Term", shape = "Ontology") +
  theme(plot.title = element_text(face = "bold"))
```




