---
title: "miRNA interactions"
author: "Jill Ashey"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(topGO)
library(rrvgo)
```

### Apoc

#### Binding summary - mRNA

Read in miranda data 
```{r}
miranda_apoc <- read.delim("../output/miRNA_interactions/apoc/miranda_strict_mRNA_parsed_apoc.txt", header = F)
colnames(miranda_apoc) <- c("miRNA", "mRNA", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

# Format miranda df 
miranda_apoc$miRNA <- sub("^>", "", miranda_apoc$miRNA)  # Remove leading ">"

dim(miranda_apoc)
length(unique(miranda_apoc$miRNA))
length(unique(miranda_apoc$mRNA))
```

Remove Cluster_598 - it is the same as Cluster_595 and targets the same number of mRNAs. 
```{r}
miranda_apoc <- miranda_apoc %>%
  filter(!miRNA == "Cluster_598")
```

Remove any possible tRNAs - anything that does not have evm at the start. 
```{r}
miranda_apoc <- miranda_apoc[grepl("evm", miranda_apoc$mRNA), ]

dim(miranda_apoc)
length(unique(miranda_apoc$miRNA))
length(unique(miranda_apoc$mRNA))
```

Calculate number of unique mRNAs targeted by each miRNA
```{r}
miRNA_target_counts_apoc <- miranda_apoc %>%
  group_by(miRNA) %>%
  summarise(num_mRNAs_targeted = n_distinct(mRNA))

min(miRNA_target_counts_apoc$num_mRNAs_targeted)
max(miRNA_target_counts_apoc$num_mRNAs_targeted)
mean(miRNA_target_counts_apoc$num_mRNAs_targeted)
median(miRNA_target_counts_apoc$num_mRNAs_targeted)
```

Perhaps collapse Cluster_595 and Cluster_598...they are the same sequence and are targeting the same number of mRNAs. 

Count how many mRNAs are targeted by multiple miRNAs
```{r}
mRNA_targeted_counts_apoc <- miranda_apoc %>%
  group_by(mRNA) %>%
  summarise(num_miRNAs_targeting = n_distinct(miRNA))

mRNAs_multiple_miRNAs_apoc <- mRNA_targeted_counts_apoc %>%
  filter(num_miRNAs_targeting > 1) %>%
  summarise(num_mRNAs = n())
mRNAs_multiple_miRNAs_apoc
```

Calculate number of bp shared in the interaction 
```{r}
miranda_apoc <- miranda_apoc %>%
  separate(query_start_end, into = c("start", "end"), sep = " ", convert = TRUE) %>%
  mutate(interaction_length = end - start)
```

Plot shared interaction bp distribution 
```{r}
ggplot(miranda_apoc, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq <- miranda_apoc %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq$interaction_length),
                                  max(df_freq$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

#### Binding summary - 3UTRs

Read in miranda data 
```{r}
miranda_3utr_apoc <- read.delim("../output/miRNA_interactions/apoc/miranda_strict_3UTR_parsed_apoc.txt", header = F)
colnames(miranda_3utr_apoc) <- c("miRNA", "three_UTR", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

# Format miranda df 
miranda_3utr_apoc$miRNA <- sub("^>", "", miranda_3utr_apoc$miRNA)  # Remove leading ">"

dim(miranda_3utr_apoc)
length(unique(miranda_3utr_apoc$miRNA))
length(unique(miranda_3utr_apoc$three_UTR))
```

Remove Cluster_598 - it is the same as Cluster_595 and targets the same number of mRNAs. 
```{r}
miranda_3utr_apoc <- miranda_3utr_apoc %>%
  filter(!miRNA == "Cluster_598")

dim(miranda_3utr_apoc)
length(unique(miranda_3utr_apoc$miRNA))
length(unique(miranda_3utr_apoc$three_UTR))
```

In order to connect the 3UTRs with their associated gene names, I need to read in `apoc_3UTRs_closest.gff` and associate the gene names in the attribute column with the data from the three_UTR column in the `miranda_3utr_apoc` df. 

Read in the gff
```{r}
apoc_3UTR_closest <- read.delim("~/Desktop/GFFs/apoc/apoc_3UTRs_closest.gff", header = F)
```

Make a new column that will match the three_UTR column in `miranda_3utr_apoc`
```{r}
# Subtract 1 to match miranda_3utr_apoc
apoc_3UTR_closest$V4 <- apoc_3UTR_closest$V4 - 1

# Make new col
apoc_3UTR_closest$three_UTR <- with(apoc_3UTR_closest, 
                                   paste0(V3, "::", V1, ":", V4, "-", V5, "(", V7, ")"))
```

Join the two dfs 
```{r}
merged_3UTR_apoc <- left_join(miranda_3utr_apoc, apoc_3UTR_closest, by = "three_UTR")
```

Extract gene names and remove other cols
```{r}
merged_3UTR_apoc <- merged_3UTR_apoc %>%
  mutate(gene_name = str_extract(V9, "(?<=ID=)[^;]+")  ) %>%
  dplyr::select(-starts_with("V"))

# Select only unique rows
merged_3UTR_apoc <- unique(merged_3UTR_apoc)
```

Remove any possible tRNAs - anything that does not have evm at the start. 
```{r}
merged_3UTR_apoc <- merged_3UTR_apoc[grepl("evm", merged_3UTR_apoc$gene_name), ]
merged_3UTR_apoc$gene_name <- sub("TU", "model", merged_3UTR_apoc$gene_name)

dim(merged_3UTR_apoc)
length(unique(merged_3UTR_apoc$miRNA))
length(unique(merged_3UTR_apoc$three_UTR))
length(unique(merged_3UTR_apoc$gene_name))
```

Write out new csv with gene names associated with the 3UTRs.
```{r}
write.csv(merged_3UTR_apoc, "../output/miRNA_interactions/apoc/miranda_strict_3UTR_parsed_genename_apoc.csv")
```

Calculate number of unique 3UTR seqs targeted by each miRNA
```{r}
miRNA_target_counts_apoc <- merged_3UTR_apoc %>%
  group_by(miRNA) %>%
  summarise(num_mRNAs_targeted = n_distinct(three_UTR))

min(miRNA_target_counts_apoc$num_mRNAs_targeted)
max(miRNA_target_counts_apoc$num_mRNAs_targeted)
mean(miRNA_target_counts_apoc$num_mRNAs_targeted)
median(miRNA_target_counts_apoc$num_mRNAs_targeted)
```

Count how many 3UTRs are targeted by multiple miRNAs
```{r}
three_UTR_targeted_counts_apoc <- merged_3UTR_apoc %>%
  group_by(three_UTR) %>%
  summarise(num_miRNAs_targeting = n_distinct(miRNA))

three_UTR_multiple_miRNAs_apoc <- three_UTR_targeted_counts_apoc %>%
  filter(num_miRNAs_targeting > 1) %>%
  summarise(num_3UTR = n())
three_UTR_multiple_miRNAs_apoc
```

Calculate number of bp shared in the interaction 
```{r}
merged_3UTR_apoc <- merged_3UTR_apoc %>%
  separate(query_start_end, into = c("start", "end"), sep = " ", convert = TRUE) %>%
  mutate(interaction_length = end - start)
```

Plot shared interaction bp distribution 
```{r}
ggplot(merged_3UTR_apoc, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_3UTR_apoc <- merged_3UTR_apoc %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_3UTR_apoc, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq$interaction_length),
                                  max(df_freq$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

#### mRNA and 3UTR binding comparisons 

Any overlap in genes from mRNA v. 3UTR binding?
```{r}
apoc_mrna_genes <- unique(miranda_apoc$mRNA)
apoc_3utr_genes <- unique(merged_3UTR_apoc$gene_name)

# shared 
shared_genes_apoc <- intersect(apoc_mrna_genes, apoc_3utr_genes)
```

49 genes are targeted by miRNAs in both their mRNA and 3'UTR sequence. 

Are the 49 genes targeted by miRNAs in both sequences targeted by the same miRNA in both sequences?
```{r}
miranda_apoc_filt <- miranda_apoc %>%
  filter(mRNA %in% shared_genes_apoc)

merged_3UTR_apoc_filt <- merged_3UTR_apoc %>%
  filter(gene_name %in% shared_genes_apoc)

shared_targets_apoc <- full_join(miranda_apoc_filt, merged_3UTR_apoc_filt, by = c("mRNA" = "gene_name"))
shared_targets_apoc$same_miRNA <- shared_targets_apoc$miRNA.x == shared_targets_apoc$miRNA.y
```

19 genes have the same miRNA targeting their 3' UTR and mRNA sequence. 

Join dfs for plotting number of binding bp for both regions. 
```{r}
miranda_apoc_interaction <- miranda_apoc %>%
  dplyr::select(miRNA, mRNA, interaction_length) %>%
  mutate(Binding = "CDS")

merged_3UTR_apoc_interaction <- merged_3UTR_apoc %>%
  dplyr::select(miRNA, gene_name, interaction_length) %>%
  mutate(Binding = "3' UTR") %>%
  dplyr::rename("mRNA" = "gene_name")

apoc_join_interaction <- rbind(miranda_apoc_interaction, merged_3UTR_apoc_interaction)
```

Plot number of binding bp. 
```{r}
ggplot(apoc_join_interaction, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_apoc <- apoc_join_interaction %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_apoc, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq_3UTR_apoc$interaction_length),
                                  max(df_freq_3UTR_apoc$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq_3UTR_apoc$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

### Nvec

#### Binding summary - mRNA

Read in miranda data 
```{r}
miranda_nvec <- read.delim("../output/miRNA_interactions/nvec/miranda_strict_mRNA_parsed_nvec.txt", header = F)
colnames(miranda_nvec) <- c("miRNA", "mRNA", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

# Format miranda df 
miranda_nvec$miRNA <- sub("^>", "", miranda_nvec$miRNA)  # Remove leading ">"

dim(miranda_nvec)
length(unique(miranda_nvec$miRNA))
length(unique(miranda_nvec$mRNA))
```

Remove Cluster_1959 and Cluster_1961- they are the same as Cluster_1954 and targets the same number of mRNAs. They also were all classified as nve-miR-9415. 
```{r}
miranda_nvec <- miranda_nvec %>%
  filter(!miRNA == "Cluster_1959") %>%
  filter(!miRNA == "Cluster_1961") 

dim(miranda_nvec)
length(unique(miranda_nvec$miRNA))
length(unique(miranda_nvec$mRNA))
```

Calculate number of unique mRNAs targeted by each miRNA
```{r}
miRNA_target_counts_nvec <- miranda_nvec %>%
  group_by(miRNA) %>%
  summarise(num_mRNAs_targeted = n_distinct(mRNA))

min(miRNA_target_counts_nvec$num_mRNAs_targeted)
max(miRNA_target_counts_nvec$num_mRNAs_targeted)
mean(miRNA_target_counts_nvec$num_mRNAs_targeted)
median(miRNA_target_counts_nvec$num_mRNAs_targeted)
```

Count how many mRNAs are targeted by multiple miRNAs
```{r}
mRNA_targeted_counts_nvec <- miranda_nvec %>%
  group_by(mRNA) %>%
  summarise(num_miRNAs_targeting = n_distinct(miRNA))

mRNAs_multiple_miRNAs_nvec <- mRNA_targeted_counts_nvec %>%
  filter(num_miRNAs_targeting > 1) %>%
  summarise(num_mRNAs = n())
mRNAs_multiple_miRNAs_nvec
```

Calculate number of bp shared in the interaction 
```{r}
miranda_nvec <- miranda_nvec %>%
  separate(query_start_end, into = c("start", "end"), sep = " ", convert = TRUE) %>%
  mutate(interaction_length = end - start)
```

Plot shared interaction bp distribution 
```{r}
ggplot(miranda_nvec, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_nvec <- miranda_nvec %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_nvec, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq_nvec$interaction_length),
                                  max(df_freq_nvec$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq_nvec$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

#### Binding summary - 3UTRs

Read in miranda data 
```{r}
miranda_3utr_nvec <- read.delim("../output/miRNA_interactions/nvec/miranda_strict_3UTR_parsed_nvec.txt", header = F)
colnames(miranda_3utr_nvec) <- c("miRNA", "three_UTR", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

# Format miranda df 
miranda_3utr_nvec$miRNA <- sub("^>", "", miranda_3utr_nvec$miRNA)  # Remove leading ">"

dim(miranda_3utr_nvec)
length(unique(miranda_3utr_nvec$miRNA))
length(unique(miranda_3utr_nvec$three_UTR))
```

Remove Cluster_1959 and Cluster_1961- they are the same as Cluster_1954 and targets the same number of mRNAs. They also were all classified as nve-miR-9415. 
```{r}
miranda_3utr_nvec <- miranda_3utr_nvec %>%
  filter(!miRNA == "Cluster_1959") %>%
  filter(!miRNA == "Cluster_1961") 

dim(miranda_3utr_nvec)
length(unique(miranda_3utr_nvec$miRNA))
length(unique(miranda_3utr_nvec$three_UTR))
```

In order to connect the 3UTRs with their associated gene names, I need to read in `nvec.GFFannotation.3UTR_1kb_corrected.gff` and associate the gene names in the attribute column with the data from the three_UTR column in the `miranda_3utr_nvec` df. 

Read in the gff
```{r}
nvec_3UTR <- read.delim("~/Desktop/GFFs/nvec/nvec.GFFannotation.3UTR_1kb_corrected.gff", header = F)
```

Make a new column that will match the three_UTR column in `miranda_3utr_apoc`
```{r}
# Subtract 1 to match miranda_3utr_apoc
nvec_3UTR$V4 <- nvec_3UTR$V4 - 1

# Make new col
nvec_3UTR$three_UTR <- with(nvec_3UTR, 
                                   paste0(V3, "::", V1, ":", V4, "-", V5, "(", V7, ")"))
```

Join the two dfs 
```{r}
merged_3UTR_nvec <- left_join(miranda_3utr_nvec, nvec_3UTR, by = "three_UTR")
```

Extract gene names and remove other cols
```{r}
merged_3UTR_nvec <- merged_3UTR_nvec %>%
  mutate(gene_name = str_extract(V9, "(?<=ID=)[^;]+")  ) %>%
  dplyr::select(-starts_with("V"))

# Select only unique rows
merged_3UTR_nvec <- unique(merged_3UTR_nvec)

dim(merged_3UTR_nvec)
length(unique(merged_3UTR_nvec$miRNA))
length(unique(merged_3UTR_nvec$three_UTR))
```

Write out new csv with gene names associated with the 3UTRs.
```{r}
write.csv(merged_3UTR_nvec, "../output/miRNA_interactions/nvec/miranda_strict_3UTR_parsed_genename_nvec.csv")
```

Calculate number of unique 3UTR seqs targeted by each miRNA
```{r}
miRNA_target_counts_nvec <- merged_3UTR_nvec %>%
  group_by(miRNA) %>%
  summarise(num_mRNAs_targeted = n_distinct(three_UTR))

min(miRNA_target_counts_nvec$num_mRNAs_targeted)
max(miRNA_target_counts_nvec$num_mRNAs_targeted)
mean(miRNA_target_counts_nvec$num_mRNAs_targeted)
median(miRNA_target_counts_nvec$num_mRNAs_targeted)
```

Count how many 3UTRs are targeted by multiple miRNAs
```{r}
three_UTR_targeted_counts_nvec <- merged_3UTR_nvec %>%
  group_by(three_UTR) %>%
  summarise(num_miRNAs_targeting = n_distinct(miRNA))

three_UTR_multiple_miRNAs_nvec <- three_UTR_targeted_counts_nvec %>%
  filter(num_miRNAs_targeting > 1) %>%
  summarise(num_3UTR = n())
three_UTR_multiple_miRNAs_nvec
```

Calculate number of bp shared in the interaction 
```{r}
merged_3UTR_nvec <- merged_3UTR_nvec %>%
  separate(query_start_end, into = c("start", "end"), sep = " ", convert = TRUE) %>%
  mutate(interaction_length = end - start)
```

Plot shared interaction bp distribution 
```{r}
ggplot(merged_3UTR_nvec, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_3UTR_nvec <- merged_3UTR_nvec %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_3UTR_nvec, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq_3UTR_nvec$interaction_length),
                                  max(df_freq_3UTR_nvec$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq_3UTR_nvec$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

#### mRNA and 3UTR binding comparisons 

Any overlap in genes from mRNA v. 3UTR binding?
```{r}
nvec_mrna_genes <- unique(miranda_nvec$mRNA)
nvec_3utr_genes <- unique(merged_3UTR_nvec$gene_name)

# shared 
shared_genes_nvec <- intersect(nvec_mrna_genes, nvec_3utr_genes)
```

134 genes are targeted by miRNAs in both their mRNA and 3'UTR sequence. 

Are the 134 genes targeted by miRNAs in both sequences targeted by the same miRNA in both sequences?
```{r}
miranda_nvec_filt <- miranda_nvec %>%
  filter(mRNA %in% shared_genes_nvec)

merged_3UTR_nvec_filt <- merged_3UTR_nvec %>%
  filter(gene_name %in% shared_genes_nvec)

shared_targets_nvec <- full_join(miranda_nvec_filt, merged_3UTR_nvec_filt, by = c("mRNA" = "gene_name"))
shared_targets_nvec$same_miRNA <- shared_targets_nvec$miRNA.x == shared_targets_nvec$miRNA.y
```

28 genes have the same miRNA targeting their 3' UTR and mRNA sequence. 

Join dfs for plotting number of binding bp for both regions. 
```{r}
miranda_nvec_interaction <- miranda_nvec %>%
  dplyr::select(miRNA, mRNA, interaction_length) %>%
  mutate(Binding = "CDS")

merged_3UTR_nvec_interaction <- merged_3UTR_nvec %>%
  dplyr::select(miRNA, gene_name, interaction_length) %>%
  mutate(Binding = "3' UTR") %>%
  dplyr::rename("mRNA" = "gene_name")

join_interaction_nvec <- rbind(miranda_nvec_interaction, merged_3UTR_nvec_interaction)
```

Plot number of binding bp. 
```{r}
ggplot(join_interaction_nvec, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_nvec <- join_interaction_nvec %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_nvec, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq_nvec$interaction_length),
                                  max(df_freq_nvec$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq_3UTR_nvec$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

## Ahya

#### Binding summary - mRNA

Read in miranda data 
```{r}
miranda_ahya <- read.delim("../output/miRNA_interactions/ahya/miranda_strict_mRNA_parsed_ahya.txt", header = F)
colnames(miranda_ahya) <- c("miRNA", "mRNA", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

# Format miranda df 
miranda_ahya$miRNA <- sub("^>", "", miranda_ahya$miRNA)  # Remove leading ">"
miranda_ahya$mRNA <- sub("-RA", "", miranda_ahya$mRNA)

dim(miranda_ahya)
length(unique(miranda_ahya$miRNA))
length(unique(miranda_ahya$mRNA))
```

Calculate number of unique mRNAs targeted by each miRNA
```{r}
miRNA_target_counts_ahya <- miranda_ahya %>%
  group_by(miRNA) %>%
  summarise(num_mRNAs_targeted = n_distinct(mRNA))

min(miRNA_target_counts_ahya$num_mRNAs_targeted)
max(miRNA_target_counts_ahya$num_mRNAs_targeted)
mean(miRNA_target_counts_ahya$num_mRNAs_targeted)
median(miRNA_target_counts_ahya$num_mRNAs_targeted)
```

Count how many mRNAs are targeted by multiple miRNAs
```{r}
mRNA_targeted_counts_ahya <- miranda_ahya %>%
  group_by(mRNA) %>%
  summarise(num_miRNAs_targeting = n_distinct(miRNA))

mRNA_targeted_counts_ahya <- mRNA_targeted_counts_ahya %>%
  filter(num_miRNAs_targeting > 1) %>%
  summarise(num_mRNAs = n())
mRNA_targeted_counts_ahya
```

Calculate number of bp shared in the interaction 
```{r}
miranda_ahya <- miranda_ahya %>%
  separate(query_start_end, into = c("start", "end"), sep = " ", convert = TRUE) %>%
  mutate(interaction_length = end - start)
```

Plot shared interaction bp distribution 
```{r}
ggplot(miranda_ahya, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_ahya <- miranda_ahya %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_ahya, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq_ahya$interaction_length),
                                  max(df_freq_ahya$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq_ahya$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

#### Binding summary - 3UTRs

Read in miranda data 
```{r}
miranda_3utr_ahya <- read.delim("../output/miRNA_interactions/ahya/miranda_strict_3UTR_parsed_ahya.txt", header = F)
colnames(miranda_3utr_ahya) <- c("miRNA", "three_UTR", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

# Format miranda df 
miranda_3utr_ahya$miRNA <- sub("^>", "", miranda_3utr_ahya$miRNA)  # Remove leading ">"

dim(miranda_3utr_ahya)
length(unique(miranda_3utr_ahya$miRNA))
length(unique(miranda_3utr_ahya$three_UTR))
```

In order to connect the 3UTRs with their associated gene names, I need to read in `ahya.GFFannotation.3UTR_1kb_corrected.gff` and associate the gene names in the attribute column with the data from the three_UTR column in the `miranda_3utr_ahya` df. 

Read in the gff
```{r}
ahya_3UTR <- read.delim("~/Desktop/GFFs/ahya/ahya.GFFannotation.3UTR_1kb_corrected.gff", header = F)
```

Make a new column that will match the three_UTR column in `miranda_3utr_apoc`
```{r}
# Subtract 1 to match miranda_3utr_apoc
ahya_3UTR$V4 <- ahya_3UTR$V4 - 1

# Make new col
ahya_3UTR$three_UTR <- with(ahya_3UTR, 
                                   paste0(V3, "::", V1, ":", V4, "-", V5, "(", V7, ")"))
```

Join the two dfs 
```{r}
merged_3UTR_ahya <- left_join(miranda_3utr_ahya, ahya_3UTR, by = "three_UTR")
```

Extract gene names and remove other cols
```{r}
merged_3UTR_ahya <- merged_3UTR_ahya %>%
  mutate(gene_name = str_extract(V9, "(?<=ID=)[^;]+")  ) %>%
  dplyr::select(-starts_with("V"))

# Select only unique rows and remove -RA
merged_3UTR_ahya <- unique(merged_3UTR_ahya)
merged_3UTR_ahya$gene_name <- gsub("-RA", "", merged_3UTR_ahya$gene_name)

dim(merged_3UTR_ahya)
length(unique(merged_3UTR_ahya$miRNA))
length(unique(merged_3UTR_ahya$three_UTR))
```

Write out new csv with gene names associated with the 3UTRs.
```{r}
write.csv(merged_3UTR_ahya, "../output/miRNA_interactions/ahya/miranda_strict_3UTR_parsed_genename_ahya.csv")
```

Calculate number of bp shared in the interaction 
```{r}
merged_3UTR_ahya <- merged_3UTR_ahya %>%
  separate(query_start_end, into = c("start", "end"), sep = " ", convert = TRUE) %>%
  mutate(interaction_length = end - start)
```

Plot shared interaction bp distribution 
```{r}
ggplot(merged_3UTR_ahya, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_ahya <- merged_3UTR_ahya %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_ahya, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq_ahya$interaction_length),
                                  max(df_freq_ahya$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq_ahya$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

#### mRNA and 3UTR binding comparisons 

Any overlap in genes from mRNA v. 3UTR binding?
```{r}
ahya_mrna_genes <- unique(miranda_ahya$mRNA)
ahya_3utr_genes <- unique(merged_3UTR_ahya$gene_name)

# shared 
shared_genes_ahya <- intersect(ahya_mrna_genes, ahya_3utr_genes)
```

359 genes are targeted by miRNAs in both their mRNA and 3'UTR sequence. 

Are the 359 genes targeted by miRNAs in both sequences targeted by the same miRNA in both sequences?
```{r}
miranda_ahya_filt <- miranda_ahya %>%
  filter(mRNA %in% shared_genes_ahya)

merged_3UTR_ahya_filt <- merged_3UTR_ahya %>%
  filter(gene_name %in% shared_genes_ahya)

shared_targets_ahya <- full_join(miranda_ahya_filt, merged_3UTR_ahya_filt, by = c("mRNA" = "gene_name"))
shared_targets_ahya$same_miRNA <- shared_targets_ahya$miRNA.x == shared_targets_ahya$miRNA.y
```

Join dfs for plotting number of binding bp for both regions. 
```{r}
miranda_ahya_interaction <- miranda_ahya %>%
  dplyr::select(miRNA, mRNA, interaction_length) %>%
  mutate(Binding = "CDS")

merged_3UTR_ahya_interaction <- merged_3UTR_ahya %>%
  dplyr::select(miRNA, gene_name, interaction_length) %>%
  mutate(Binding = "3' UTR") %>%
  dplyr::rename("mRNA" = "gene_name")

join_interaction_ahya <- rbind(miranda_ahya_interaction, merged_3UTR_ahya_interaction)
```

Plot number of binding bp. 
```{r}
ggplot(join_interaction_ahya, aes(x = interaction_length)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    binwidth = 1, fill = "steelblue", color = "black"
  ) +
  labs(
    x = "Total bp shared",
    y = "Frequency")

# calculate percentages per bin
df_freq_ahya <- join_interaction_ahya %>%
  count(interaction_length) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(df_freq_ahya, aes(x = interaction_length, y = perc)) +
  # filled frequency curve
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 1.2) +
  
  # axis labels & title
  labs(
    x = "Total bp shared",
    y = "Frequency (%)"
  ) +
  
  # axis breaks
  scale_x_continuous(breaks = seq(min(df_freq_ahya$interaction_length),
                                  max(df_freq_ahya$interaction_length), by = 2)) +
  scale_y_continuous(breaks = seq(0, max(df_freq_ahya$perc), by = 5)) +
  
  # clean minimal theme
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```


## Species comparisons 

Plot binding percentages across mRNA, 3' UTR or both. 
```{r}
# Data frame with counts
percent <- data.frame(
  Species = rep(c("Acropora hyacinthus", "Astrangia poculata", "Nematostella vectensis"), each = 3),
  Binding = rep(c("CDS only", "3' UTR only", "Both"), times = 3),
  Count = c(3852, 1991, 359,
            2419, 826, 49,
            5806, 446, 134))

# Calculate percentages within each species
percent <- percent %>%
  group_by(Species) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

# Calculate total counts per species for labels
totals <- percent %>%
  group_by(Species) %>%
  summarize(Total = sum(Count))

# Plot stacked bar plot with total counts on top of each bar
feature_plot <- ggplot(percent, aes(x = Species, y = Percent, fill = Binding)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", position = position_stack(reverse = F)) +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  # geom_text(
  #   data = totals,
  #   aes(x = Species, y = 105, label = Total),
  #   inherit.aes = FALSE,   
  #   size = 4) +
  scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 100, by = 10),
        limits = c(0, 105)) +
  ylab("Percentage (%)") +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  coord_cartesian(clip = "off") +
  ylab("Percentage (%)") +
  xlab("") +
  theme_classic() + 
  theme(axis.text = element_text(color = "black", size = 21),
    axis.title = element_text(size = 23, face = "bold"),
    axis.text.x = element_text(face = "italic"),
    legend.title = element_text(size = 20, face = "bold"),
    legend.text = element_text(size = 19),
    legend.key.size = unit(0.6, "cm")) +
  guides(fill = guide_legend(reverse = F)); feature_plot

ggsave("../output/miRNA_interactions/shared/target_binding_feature.pdf", plot = feature_plot, width = 10, height = 8, dpi = 300)
ggsave("../output/miRNA_interactions/shared/target_binding_feature.png", plot = feature_plot, width = 10, height = 8, dpi = 300)
```

Compare number of binding bp 
```{r}
df_freq_ahya$Species <- "Acropora hyacinthus"
df_freq_apoc$Species <- "Astrangia poculata"
df_freq_nvec$Species <- "Nematostella vectensis"

df_freq_all <- rbind(df_freq_ahya, df_freq_apoc, df_freq_nvec)

# Plot
bp_plot <- ggplot(df_freq_all, aes(x = interaction_length, y = perc, fill = Species, color = Species)) +
  geom_area(alpha = 0.4, position = "identity") +
  geom_line(size = 1.1) +
  scale_fill_manual(values = c(
    "Acropora hyacinthus" = "#E69F00", 
    "Astrangia poculata"  = "#56B4E9", 
    "Nematostella vectensis" = "#009E73")) +
  scale_color_manual(values = c(
    "Acropora hyacinthus" = "#E69F00", 
    "Astrangia poculata"  = "#56B4E9", 
    "Nematostella vectensis" = "#009E73")) +
  labs(
    x = "Total bp shared",
    y = "Frequency (%)") +
  scale_y_continuous(breaks = seq(0, max(df_freq_all$perc) + 5, by = 5), 
                     expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(7, 21, by = 2)) +
  
  theme_classic() +
  theme(
    axis.title = element_text(size = 23, face = "bold"),
    axis.text = element_text(size = 21, color = "black"),
    legend.position = c(0.15, 0.95),       # Places it near the top right (95% right, 95% up)
    legend.justification = c("left", "top"), # Anchors the legend box by its top-right corner
    legend.background = element_blank(),   # Optional: removes the white box background
    legend.text = element_text(size = 15, face = "italic"), # Keeps the names italicized
    legend.title = element_text(size = 16, face = "bold"),
    plot.margin = margin(10, 10, 10, 10)
  ); bp_plot

ggsave("../output/miRNA_interactions/shared/bp_target_binding.pdf", plot = bp_plot, width = 10, height = 6, dpi = 300)
ggsave("../output/miRNA_interactions/shared/bp_target_binding.png", plot = bp_plot, width = 10, height = 6, dpi = 300)
```

Put plots together 
```{r}
combined_plot <- plot_grid(
  bp_plot, 
  feature_plot, 
  ncol = 2, 
  labels = c("A", "B"), 
  label_size = 40,
  label_x = c(0, -0.05),
  label_y = c(1, 1.02),  
  rel_widths = c(1.2, 0.8)); combined_plot

ggsave("../output/miRNA_interactions/shared/miRNA_features.pdf",
       combined_plot, width = 20, height = 8, units = "in", dpi = 600, bg = "white")
ggsave("../output/miRNA_interactions/shared/miRNA_features.png",
       combined_plot, width = 20, height = 8, units = "in", dpi = 600, bg = "white")
```



