---
title: "TE in piRNA clusters"
author: "Jill Ashey"
date: "2025-11-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(GenomicRanges)
library(MetBrewer)
library(cowplot)
```

## Apoc 

### piRNAs 

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/apoc/Apoc_piRNAs_TEs_repeats_intersect.txt")

# Isolate repeat type 
pirna_bed <- pirna_bed %>%
  mutate(repeat_type = str_extract(V13, "(?<=Motif:)[^ ]+"))
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```
2575 piRNAs overlap with TEs 

Read in repeat out file 
```{r}
repeat_map <- read.table("../output/piRNAs/Apoc/apoculata.genome.fasta.out", header = F, skip = 3, fill = T)

# Assign proper column names
colnames(repeat_map) <- c(
  "SW_score", "perc_div", "perc_del", "perc_ins", 
  "query_seq", "pos_begin", "pos_end", "pos_left", 
  "strand", "Repeat_Family", "Repeat_Class", 
  "rep_begin", "rep_end", "rep_left", "ID"
)

# Keep only the distinct names 
rep_lookup <- repeat_map %>%
  dplyr::select(Repeat_Family, Repeat_Class) %>%
  distinct()
```

Merge the intersection file and the repeat look up file 
```{r}
pirna_summary <- pirna_bed %>%
  mutate(piRNA_ID = paste(V1, V2, V3, sep = "_")) %>%
  left_join(rep_lookup, by = c("repeat_type" = "Repeat_Family")) %>%
  distinct(piRNA_ID, repeat_type, .keep_all = TRUE)
```

Reassign repeat names 
```{r}
apoc_repeat_counts <- pirna_summary %>%
  count(Repeat_Class, name = "piRNA_Count") %>%
  mutate(
    Species = "Astrangia poculata",
    Clean_Group = case_when(
      str_detect(Repeat_Class, "^DNA|^RC")                          ~ "DNA Transposon",
      str_detect(Repeat_Class, "^LTR")                              ~ "LTR Retrotransposon",
      str_detect(Repeat_Class, "^LINE")                             ~ "LINE",
      str_detect(Repeat_Class, "^SINE")                             ~ "SINE",
      Repeat_Class %in% c("tRNA", "rRNA", "snRNA")                  ~ "ncRNA",
      Repeat_Class %in% c("Simple_repeat", "Low_complexity")        ~ "Simple / Low Complexity",
      Repeat_Class == "Unknown" | is.na(Repeat_Class)              ~ "Unknown",
      TRUE                                                          ~ "Other")) %>%
  group_by(Species, Clean_Group) %>%
  summarise(piRNA_Count = sum(piRNA_Count), .groups = "drop")
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/apoc/apoc_piRNAs_genes_intersect.txt", sep = "\t")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```
3179 piRNAs overlap with genes 

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

1391 piRNAs overlap both genes and TEs.

To summarize: 
- 5537 piRNAs in Apoc
- 1184 piRNAs overlap with only TEs (2575 - 1391 = 1184)
- 1788 piRNAs overlap with only genes (3179 - 1391 = 1788)
- 1391 piRNAs overlap with both TE and genes 
- 1184+1788+1391 = 4363 piRNAs overlapping TEs, genes or both 
- 1174 piRNAs in intergenic space (5537 - 4363 = 1174)

## Nvec 

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/nvec/Nvec_piRNAs_TEs_repeats_intersect.txt")

# Isolate repeat type 
pirna_bed <- pirna_bed %>%
  mutate(repeat_type = str_extract(V13, "(?<=Motif:)[^ ]+"))
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```
13655 piRNAs overlap with TEs

Read in repeat out file 
```{r}
repeat_map <- read.table("../output/piRNAs/nvec/Nvec200.fasta.out", header = F, skip = 3, fill = T)

# Assign proper column names
colnames(repeat_map) <- c(
  "SW_score", "perc_div", "perc_del", "perc_ins", 
  "query_seq", "pos_begin", "pos_end", "pos_left", 
  "strand", "Repeat_Family", "Repeat_Class", 
  "rep_begin", "rep_end", "rep_left", "ID"
)

# Keep only the distinct names 
rep_lookup <- repeat_map %>%
  dplyr::select(Repeat_Family, Repeat_Class) %>%
  distinct()
```

Merge the intersection file and the repeat look up file 
```{r}
pirna_summary <- pirna_bed %>%
  mutate(piRNA_ID = paste(V1, V2, V3, sep = "_")) %>%
  left_join(rep_lookup, by = c("repeat_type" = "Repeat_Family")) %>%
  distinct(piRNA_ID, repeat_type, .keep_all = TRUE)
```

Reassign repeat names 
```{r}
nvec_repeat_counts <- pirna_summary %>%
  count(Repeat_Class, name = "piRNA_Count") %>%
  mutate(
    Species = "Nematostella vectensis",
    Clean_Group = case_when(
      str_detect(Repeat_Class, "^DNA|^RC")                          ~ "DNA Transposon",
      str_detect(Repeat_Class, "^LTR")                              ~ "LTR Retrotransposon",
      str_detect(Repeat_Class, "^LINE")                             ~ "LINE",
      str_detect(Repeat_Class, "^SINE")                             ~ "SINE",
      Repeat_Class %in% c("tRNA", "rRNA", "snRNA")                  ~ "ncRNA",
      Repeat_Class %in% c("Simple_repeat", "Low_complexity")        ~ "Simple / Low Complexity",
      Repeat_Class == "Unknown" | is.na(Repeat_Class)              ~ "Unknown",
      TRUE                                                          ~ "Other")) %>%
  group_by(Species, Clean_Group) %>%
  summarise(piRNA_Count = sum(piRNA_Count), .groups = "drop")
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/nvec/nvec_piRNAs_genes_intersect.txt", sep = "\t")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```
12963 piRNAs overlap with genes wc -l 

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

5479 piRNAs overlap both genes and TEs. 

To summarize: 
- 30471 piRNAs in Nvec
- 8176 piRNAs overlap with only TEs (13655 - 5479 = 8176)
- 7484 piRNAs overlap with only genes (12963 - 5479 = 7484)
- 5479 piRNAs overlap with both TE and genes 
- 8176+7484+5479 = 21139 piRNAs overlapping TEs, genes or both 
- 9332 piRNAs in intergenic space (30471 - 21139 = 9332)

## Ahya 

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/ahya/Ahya_piRNAs_TEs_repeats_intersect.txt")

# Isolate repeat type 
pirna_bed <- pirna_bed %>%
  mutate(repeat_type = str_extract(V13, "(?<=Motif:)[^ ]+"))
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```
285871 piRNAs overlap with TEs 

Read in repeat out file 
```{r}
repeat_map <- read.table("../output/piRNAs/ahya/Ahyacinthus.chrsV1.fasta.out", header = F, skip = 3, fill = T)

# Assign proper column names
colnames(repeat_map) <- c(
  "SW_score", "perc_div", "perc_del", "perc_ins", 
  "query_seq", "pos_begin", "pos_end", "pos_left", 
  "strand", "Repeat_Family", "Repeat_Class", 
  "rep_begin", "rep_end", "rep_left", "ID"
)

# Keep only the distinct names 
rep_lookup <- repeat_map %>%
  dplyr::select(Repeat_Family, Repeat_Class) %>%
  distinct()
```

Merge the intersection file and the repeat look up file 
```{r}
pirna_summary <- pirna_bed %>%
  mutate(piRNA_ID = paste(V1, V2, V3, sep = "_")) %>%
  left_join(rep_lookup, by = c("repeat_type" = "Repeat_Family")) %>%
  distinct(piRNA_ID, repeat_type, .keep_all = TRUE)
```

Reassign repeat names 
```{r}
ahya_repeat_counts <- pirna_summary %>%
  count(Repeat_Class, name = "piRNA_Count") %>%
  mutate(
    Species = "Acropora hyacinthus",
    Clean_Group = case_when(
      str_detect(Repeat_Class, "^DNA|^RC")                          ~ "DNA Transposon",
      str_detect(Repeat_Class, "^LTR")                              ~ "LTR Retrotransposon",
      str_detect(Repeat_Class, "^LINE")                             ~ "LINE",
      str_detect(Repeat_Class, "^SINE")                             ~ "SINE",
      Repeat_Class %in% c("tRNA", "rRNA", "snRNA")                  ~ "ncRNA",
      Repeat_Class %in% c("Simple_repeat", "Low_complexity")        ~ "Simple / Low Complexity",
      Repeat_Class == "Unknown" | is.na(Repeat_Class)              ~ "Unknown",
      TRUE                                                          ~ "Other")) %>%
  group_by(Species, Clean_Group) %>%
  summarise(piRNA_Count = sum(piRNA_Count), .groups = "drop")
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/ahya/Ahya_piRNAs_genes_intersect.txt")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```
119663 piRNAs overlap with genes 

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

84075 piRNAs overlap with both TEs and genes. 

To summarize: 
- 364493 piRNAs in Ahya
- 201796 piRNAs overlap with only TE (285871 - 84075 = 201796)
- 35588 piRNAs overlap with only genes (119663 - 84075 = 35588)
- 84075 piRNAs overlap with both TE and genes 
- 201796+35588+84075 = 321459 piRNAs overlapping TEs, genes or both 
- 43034 piRNAs in intergenic space (364493 - 321459 = 43034)

## Species comparisons

Plot TE overlap types across spp 
```{r}
all_repeat <- rbind(ahya_repeat_counts, apoc_repeat_counts, nvec_repeat_counts)

# Standardize categories across both species
all_repeat_clean <- all_repeat %>%
  mutate(Clean_Group = case_when(
    Clean_Group %in% c("Low Complexity", "Simple Repeat", "Simple / Low Complexity") ~ "Simple / Low Complexity",
    Clean_Group == "Rolling Circle (Helitron)" ~ "DNA Transposon",
    TRUE ~ Clean_Group)) %>%
  group_by(Species, Clean_Group) %>%
  summarise(piRNA_Count = sum(piRNA_Count), .groups = "drop")

# Calculate percentages (0 to 100) per species
percent <- all_repeat_clean %>%
  group_by(Species) %>%
  mutate(Percent = (piRNA_Count / sum(piRNA_Count)) * 100) %>%
  ungroup() %>%
  dplyr::rename(Element = Clean_Group)

# Set factor levels for consistent legend & stacking order
percent$Element <- factor(
  percent$Element,
  levels = c(
    "DNA Transposon",
    "LINE",
    "LTR Retrotransposon",
    "SINE",
    "Simple / Low Complexity",
    "ncRNA",
    "Unknown"))

# Colors
earthy_forest <- c(
  "DNA Transposon"          = "#2D5A27", # Deep Pine Green
  "LINE"                    = "#4E9F3D", # Bright Emerald Green
  "LTR Retrotransposon"     = "#8FBC8F", # Soft Sage Green
  "SINE"                    = "#D97706", # Amber / Gold
  "Simple / Low Complexity" = "#9CA3AF", # Muted Cool Gray
  "ncRNA"                   = "#C05621", # Terracotta / Warm Rust
  "Unknown"                 = "#E5E7EB"  # Very Light Gray
)
botanical_greens <- c(
  "DNA Transposon"          = "#1E4620", # Very Dark Forest
  "LINE"                    = "#3B7A57", # Deep Sea Green / Mint
  "LTR Retrotransposon"     = "#70A288", # Soft Botanical Green
  "SINE"                    = "#A3C9A8", # Pale Sage
  "Simple / Low Complexity" = "#84A98C", # Slate Green
  "ncRNA"                   = "#D8F3DC", # Mint Ice
  "Unknown"                 = "#E0E0E0"  # Neutral Light Gray
)
algae_ocean <- c(
  "DNA Transposon"          = "#0B4B32", # Deep Teal-Green
  "LINE"                    = "#2D8A4E", # Mid Emerald
  "LTR Retrotransposon"     = "#74C476", # Spring Green
  "SINE"                    = "#C7E9C0", # Soft Yellow-Green
  "Simple / Low Complexity" = "#41B6C4", # Ocean Blue-Green
  "ncRNA"                   = "#FED976", # Soft Warm Yellow
  "Unknown"                 = "#E0E0E0"  # Neutral Light Gray
)

# Plot
pirna_repeat_overlap_spp <- ggplot(percent, aes(x = Species, y = Percent, fill = Element)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", position = position_stack(reverse = TRUE)) +
  #scale_fill_brewer(palette = "Demuth", direction = -1) +
  #scale_fill_viridis_d(option = "turbo", direction = -1) +
  scale_fill_manual(values = botanical_greens, name = "Repetitive Element") +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = seq(0, 100, by = 10),
    limits = c(0, 105)) +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  coord_cartesian(clip = "off") +
  ylab("Percentage (%)") +
  xlab("") +
  theme_classic() + 
  theme(
    axis.text = element_text(color = "black", size = 20),
    axis.title = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(face = "italic"),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 15),
    legend.key.size = unit(0.6, "cm")) +
  guides(fill = guide_legend(reverse = TRUE)); pirna_repeat_overlap_spp

ggsave("../output/piRNAs/cross-species/piRNA_repeat_overlaps.pdf",
       pirna_repeat_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/cross-species/piRNA_repeat_overlaps.png",
       pirna_repeat_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
```

Plot percent of piRNAs overlapping with TEs, genes, both or intergenic regions 
```{r}
# Data frame with counts
percent <- data.frame(
  Species = rep(c("Acropora hyacinthus", "Astrangia poculata", "Nematostella vectensis"), each = 4),
  Element = rep(c("TEs only", "Genes only", "Both", "Intergenic"), times = 3),
  Count = c(201796, 35588, 84075, 43034, 
            1184, 1788, 1391, 1174,
            8176, 7484, 5479, 9332))

# Calculate percentages within each species
percent <- percent %>%
  group_by(Species) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

# Ensure the order of factors is correct
percent$Element <- factor(percent$Element, 
                                levels = c("Intergenic", "Genes only", "Both", "TEs only"))

pirna_overlap_spp <- ggplot(percent, aes(x = Species, y = Percent, fill = Element)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", position = position_stack(reverse = T)) +
  scale_fill_brewer(palette = "Greens", direction = -1, name = "Genomic Element") +
  scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 100, by = 10),
        limits = c(0, 105)) +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  coord_cartesian(clip = "off") +
  ylab("Percentage (%)") +
  xlab("") +
  theme_classic() + 
  theme(axis.text = element_text(color = "black", size = 20),
    axis.title = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(face = "italic"),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 15),
    legend.key.size = unit(0.6, "cm")) +
  guides(fill = guide_legend(reverse = TRUE)); pirna_overlap_spp

ggsave("../output/piRNAs/cross-species/piRNA_overlaps.pdf",
       pirna_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/cross-species/piRNA_overlaps.png",
       pirna_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
```

## Ping pong signatures for piRNAs 

Set custom plot theme
```{r}
theme_custom <- theme_classic(base_size = 9) +
  theme(
    text = element_text(family = "sans"),
    axis.text = element_text(color = "black", size = 20),
    axis.title = element_text(size = 22, face = "bold"),
    strip.text = element_text(face = "bold", size = 20),
    plot.title = element_text(face = "italic", size = 24, hjust = 0.5),
    #panel.border = element_rect(color = "grey60", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    #panel.grid.major.x = element_line(color = "grey90", linewidth = 0.3),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.4, "cm"))
    #plot.margin = margin(2,2,2,2))
```

### Nvec 

Ping pong signatures - overlap at position 10 

Read in data 
```{r}
pp_nvec <- read.csv("../output/piRNAs/nvec/piRNA_pingpong_nvec.csv")
```

Calculate mean and se per overlap across reps
```{r}
summary_data_nvec <- pp_nvec %>%
  group_by(overlap_bp) %>%
  summarise(
    mean_read_pairs = mean(read_pairs, na.rm = TRUE),
    sd_read_pairs = sd(read_pairs, na.rm = TRUE),
    n = sum(!is.na(read_pairs)),  # count non-NA values per group
    sem_read_pairs = sd_read_pairs / sqrt(n))
```

Filter so we only plot up to 20bp
```{r}
summary_data_filt_nvec <- summary_data_nvec %>%
  filter(overlap_bp <= 20)
```

Average z-scores across the reps
```{r}
zscore_nvec <- mean(
  5.42328080965236, # nvec 1
  8.81178350311666, # nvec 2
  13.099921789492, # nvec 3
  10.001391045776 # nvec 4
)
print(zscore_nvec)
```

Plot
```{r}
overlap_plot_nvec <- ggplot(summary_data_filt_nvec, aes(x = overlap_bp, y = mean_read_pairs)) +
  geom_col(fill = "forestgreen", color = "black", width = 0.7) +  # black outline and narrower bars
  geom_errorbar(aes(ymin = mean_read_pairs - sem_read_pairs,
                    ymax = mean_read_pairs + sem_read_pairs),
                width = 0.2) +
  labs(x = "5' Overlap (nt)", y = "Read pairs score", title = "Nematostella vectensis") +
  geom_label(
    aes(x = 12, y = 55, label = "Z-score = 5.42"),
    fill = "white",       # background color of the box
    color = "black",      # text color
    label.size = 0.5,     # thickness of the box border
    fontface = "plain",
    hjust = 0,
    vjust = 0,
    size = 6) +
  scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 80, by = 10),
        limits = c(0, 80)) +
  scale_x_continuous(
        expand = c(0, 0),
        breaks = seq(0, 20, by = 1),
        limits = c(0, 20.5)) +
  theme_custom; overlap_plot_nvec

ggsave("../output/piRNAs/nvec/ping_pong_10nt_overlap_nvec.pdf",
       overlap_plot_nvec, width = 6, height = 4, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/nvec/ping_pong_10nt_overlap_nvec.png",
       overlap_plot_nvec, width = 6, height = 4, units = "in", dpi = 600, bg = "white")
```

### Apoc 

Ping pong signatures - overlap at position 10 

Read in data 
```{r}
pp_apoc <- read.csv("../output/piRNAs/apoc/piRNA_pingpong_apoc.csv")
```

Calculate mean and se per overlap across reps
```{r}
summary_data_apoc <- pp_apoc %>%
  group_by(overlap_bp) %>%
  summarise(
    mean_read_pairs = mean(read_pairs, na.rm = TRUE),
    sd_read_pairs = sd(read_pairs, na.rm = TRUE),
    n = sum(!is.na(read_pairs)),  # count non-NA values per group
    sem_read_pairs = sd_read_pairs / sqrt(n))
```

Filter so we only plot up to 20bp
```{r}
summary_data_filt_apoc <- summary_data_apoc %>%
  filter(overlap_bp <= 20)
```

Average z-scores across the reps
```{r}
zscore_apoc <- mean(
  11.8021967062251, # apoc 2
  35.5967468197136, # apoc 3
  8.86279373084146 # apoc 4
)
print(zscore_apoc)
```

Plot
```{r}
overlap_plot_apoc <- ggplot(summary_data_filt_nvec, aes(x = overlap_bp, y = mean_read_pairs)) +
  geom_col(fill = "forestgreen", color = "black", width = 0.7) +  # black outline and narrower bars
  geom_errorbar(aes(ymin = mean_read_pairs - sem_read_pairs,
                    ymax = mean_read_pairs + sem_read_pairs),
                width = 0.2) +
  labs(x = "5' Overlap (nt)", y = "Read pairs score", title = "Astrangia poculata") +
  geom_label(
    aes(x = 12, y = 55, label = "Z-score = 11.80"),
    fill = "white",       # background color of the box
    color = "black",      # text color
    label.size = 0.5,     # thickness of the box border
    fontface = "plain",
    hjust = 0,
    vjust = 0,
    size = 6) +
  scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 80, by = 10),
        limits = c(0, 80)) +
  scale_x_continuous(
        expand = c(0, 0),
        breaks = seq(0, 20, by = 1),
        limits = c(0, 20.5)) +
    theme_custom +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank()); overlap_plot_apoc

ggsave("../output/piRNAs/apoc/ping_pong_10nt_overlap_apoc.pdf",
       overlap_plot_apoc, width = 6, height = 4, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/apoc/ping_pong_10nt_overlap_apoc.png",
       overlap_plot_apoc, width = 6, height = 4, units = "in", dpi = 600, bg = "white")
```

### Ahya 

Ping pong signatures - overlap at position 10 

Read in data 
```{r}
pp_ahya <- read.csv("../output/piRNAs/ahya/piRNA_pingpong_ahya.csv")
```

Calculate mean and se per overlap across reps
```{r}
summary_data_ahya <- pp_ahya %>%
  group_by(overlap_bp) %>%
  summarise(
    mean_read_pairs = mean(read_pairs, na.rm = TRUE),
    sd_read_pairs = sd(read_pairs, na.rm = TRUE),
    n = sum(!is.na(read_pairs)),  # count non-NA values per group
    sem_read_pairs = sd_read_pairs / sqrt(n))
```

Filter so we only plot up to 20bp
```{r}
summary_data_filt_ahya <- summary_data_ahya %>%
  filter(overlap_bp <= 20)
```

Average z-scores across the reps
```{r}
zscore_ahya <- mean(
  8.31329382842131, # ahya 1
  2.13223121260409, # ahya 2
  30.923292527817, # ahya 3
  7.64386102569554 # ahya 4
)
print(zscore_ahya)
```

Plot
```{r}
overlap_plot_ahya <- ggplot(summary_data_filt_ahya, aes(x = overlap_bp, y = mean_read_pairs)) +
  geom_col(fill = "forestgreen", color = "black", width = 0.7) +  # black outline and narrower bars
  geom_errorbar(aes(ymin = mean_read_pairs - sem_read_pairs,
                    ymax = mean_read_pairs + sem_read_pairs),
                width = 0.2) +
  labs(x = "5' Overlap (nt)", y = "Read pairs score", title = "Acropora hyacinthus") +
  geom_label(
    aes(x = 12, y = 250000, label = "Z-score = 8.31"),
    fill = "white",       # background color of the box
    color = "black",      # text color
    label.size = 0.5,     # thickness of the box border
    fontface = "plain",
    hjust = 0,
    vjust = 0,
    size = 6) +
   scale_y_continuous(
         expand = c(0, 0)) +
  scale_x_continuous(
        expand = c(0, 0),
        breaks = seq(0, 20, by = 1),
        limits = c(0, 20.5)) +
  theme_custom +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank()); overlap_plot_ahya

ggsave("../output/piRNAs/ahya/ping_pong_10nt_overlap_ahya.pdf",
       overlap_plot_ahya, width = 6, height = 4, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/ahya/ping_pong_10nt_overlap_ahya.png",
       overlap_plot_ahya, width = 6, height = 4, units = "in", dpi = 600, bg = "white")
```

## Plot all piRNA things 

```{r}
# Left col - ping pong signatures 
left_col <- plot_grid(
  overlap_plot_ahya, 
  overlap_plot_apoc, 
  overlap_plot_nvec, 
  ncol = 1, 
  align = "v",
  labels = c("A", "B", "C"), 
  label_size = 22
)

# Right col - repeat overlap and genomic overlap 
right_col <- plot_grid(
  pirna_repeat_overlap_spp, 
  pirna_overlap_spp, 
  ncol = 1, 
  align = "v",
  labels = c("D", "E"), 
  label_size = 22,
  rel_heights = c(1, 1) # Adjust height ratio if one needs more space
)

# Combine cols
full_figure <- plot_grid(
  left_col, 
  right_col, 
  ncol = 2, 
  rel_widths = c(0.8, 1.2)); full_figure

ggsave("../output/piRNAs/cross-species/piRNA_features.pdf",
       full_figure, width = 20, height = 16, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/cross-species/piRNA_features.png",
       full_figure, width = 20, height = 16, units = "in", dpi = 600, bg = "white")
```



