---
title: "TE in piRNA clusters"
author: "Jill Ashey"
date: "2025-11-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(GenomicRanges)
library(topGO)
```

## Apoc 

Read in repeatmasker out file 
```{r}
out_file <- read.table("~/Desktop/GFFs/apoc/apoculata.genome.fasta.out", comment.char = "", header = F, fill = T, skip = 2)
```

Retain only chromosome, start, stop, TE name, Family Name and percent divergence
```{r}
bed_file <- out_file[,c(5:7,11,2)]

#remove empty lines 
bed_file <- bed_file %>%
  drop_na(.)
```

Calculate the proportion of each repeat type 
```{r}
bed_file <- bed_file %>%
  mutate(length = V7 - V6 + 1)  # Calculate length of each interval

# Summarize total length per repeat type
length_per_repeat <- bed_file %>%
  group_by(V11) %>%
  summarise(total_length = sum(length)) %>%
  ungroup()

# Calculate total length of all repeats
total_length_all <- sum(length_per_repeat$total_length)

# Calculate proportion of each repeat type
length_per_repeat <- length_per_repeat %>%
  mutate(proportion = total_length / total_length_all) %>%
  mutate(proportion_precent = proportion*100)

# Assume length_per_repeat contains columns: V11 (repeat type), proportion (calculated proportion)
ggplot(length_per_repeat, aes(x = reorder(V11, -proportion_precent), y = proportion_precent, fill = V11)) +
  geom_bar(stat = "identity") +                       # Bar heights represent proportions
  #scale_y_continuous(labels = scales::percent) +     # Y-axis as percentages
  labs(x = "Repeat Element Type", y = "Proportion of Genome (%)", 
       title = "Proportion of Genomic Repeat Elements") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x labels for readability
```

Write out bed file
```{r}
write.table(bed_file, file = "~/Desktop/GFFs/apoc/Apoc_repeatmasker.bed", sep = "\t", col.names = F, row.names = F, quote = F)
```

### piRNA clusters

Read in bed overlap file
```{r}
cluster_bed <- read.table("../output/piRNAs/apoc/Apoc_piRNA_cluster_TEs_intersect.txt")
```

Calculate number of unique piRNA clusters (unique genomic intervals in V1:V3)
```{r}
unique_clusters <- cluster_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNA_clusters <- nrow(unique_clusters)
```

3 piRNA clusters overlap with some TE. That is so little!!! 

Calculate how many unique piRNA clusters overlap each repeat type (V7)
```{r}
clusters_per_repeat <- cluster_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNA_clusters)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(clusters_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNA clusters overlapping", title = "Proportion of piRNA clusters overlapping RepeatMasker elements - Apoc") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

And they only overlap simple repeats. 

Read in piRNA cluster-gene overlap 
```{r}
cluster_gene <- read.table("../output/piRNAs/apoc/Apoc_piRNA_cluster_genes_intersect.txt")
```

Calculate number of unique piRNA clusters overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_cluster_gene <- cluster_gene %>%
  distinct(V1, V2, V3)

num_unique_cluster_gene <- nrow(unique_cluster_gene)

## unique gene
unique_gene <- cluster_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/apoc/Apoc_piRNAs_TEs_intersect.txt")
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```

Calculate how many unique piRNAs overlap each repeat type (V7)
```{r}
pirna_per_repeat <- pirna_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNAs)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(pirna_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNAs overlapping", title = "Proportion of piRNAs overlapping RepeatMasker elements - Apoc") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/apoc/Apoc_piRNAs_genes_intersect.txt")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

142 piRNAs overlap with both TEs and genes. 

To summarize: 
- 5537 piRNAs in Apoc
- 201 piRNAs overlap with some TE
- 3179 piRNAs overlap with genes
- 142 piRNAs overlap with both TE and genes 
- 201+3179+142 = 3522 piRNAs overlapping TEs, genes or both 

This equals about ~63% of the piRNAs. The rest are in intergenic space (could be in other elements but these are unannotated). Much less unique piRNAs in Apoc than in Ahya. 

### piRNA gene overlap enrichment 

Perform GO enrichment for genes that are overlapping with piRNAs. 

#### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
apoc_gene2go<-read.delim("../data/annotations/Apoc_gene2go.tab", sep="\t")
```

Isolate gene names in `pirna_gene`
```{r}
pirna_gene <- pirna_gene %>%
  mutate(gene_name = str_extract(V12, "(?<=ID=)[^;]+"))
pirna_gene$gene_name <- gsub("TU", "model", pirna_gene$gene_name)
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
apoc_clust_genes <- as.character(unique(pirna_gene$gene_name))

# All genes 
apoc_all_genes <- as.character(apoc_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
apoc_GeneList <- factor(as.integer(apoc_all_genes %in% apoc_clust_genes))
names(apoc_GeneList) <- apoc_all_genes
```

Read in gene-to-go-mappings
```{r}
apoc_gene2go_topgo<-readMappings("../data/annotations/Apoc_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

#### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APOC_BP <-new("topGOdata", ontology="BP", gene2GO=apoc_gene2go_topgo, allGenes=apoc_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APOC_BP_FE <- runTest(GO_APOC_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APOC_BP_En <- GenTable(GO_APOC_BP, Fisher = GO_APOC_BP_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APOC_BP_En$Fisher<-as.numeric(GO_APOC_BP_En$Fisher)
GO_APOC_BP_En_sig<-GO_APOC_BP_En[GO_APOC_BP_En$Fisher<0.05,]
```

#### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APOC_MF <-new("topGOdata", ontology="MF", gene2GO=apoc_gene2go_topgo, allGenes=apoc_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APOC_MF_FE <- runTest(GO_APOC_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APOC_MF_En <- GenTable(GO_APOC_MF, Fisher = GO_APOC_MF_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APOC_MF_En$Fisher<-as.numeric(GO_APOC_MF_En$Fisher)
GO_APOC_MF_En_sig<-GO_APOC_MF_En[GO_APOC_MF_En$Fisher<0.05,]
```

#### Plotting enrichment 

Plot top 10 terms
```{r}
## Bind dfs 
GO_APOC_BP_En_sig$ontology <- "Biological Processes"
GO_APOC_MF_En_sig$ontology <- "Molecular Functions"
BP_MF_APOC <- rbind(GO_APOC_BP_En_sig, GO_APOC_MF_En_sig)

## Select top 10 terms for each ontology 
top10_per_ontology <- BP_MF_APOC %>%
  group_by(ontology) %>%
  slice_max(Significant, n = 10, with_ties = FALSE) %>%
  ungroup()
  
ggplot(top10_per_ontology, aes(x = reorder(Term, Significant), y = Significant, fill = Fisher)) +
  geom_bar(stat = "identity") +
  #geom_text(aes(label = gene_count), vjust = -0.5, color = "black", size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  labs(x = "GO Term",
       y = "Number of Significant Genes",
       fill = "Fisher Value") +
  theme_bw()
```

## Nvec 

Read in repeatmasker out file
```{r}
out_file <- read.table("~/Desktop/GFFs/nvec/Nvec200.fasta.out", comment.char = "", header = F, fill = T, skip = 2)
```

Retain only chromosome, start, stop, TE name, Family Name and percent divergence
```{r}
bed_file <- out_file[,c(5:7,11,2)]

#remove empty lines 
bed_file <- bed_file %>%
  drop_na(.)
```

Write out bed file
```{r}
write.table(bed_file, file = "~/Desktop/GFFs/nvec/nvec_repeatmasker.bed", sep = "\t", col.names = F, row.names = F, quote = F)
```

### piRNA clusters 

Read in bed overlap file
```{r}
cluster_bed <- read.table("../output/piRNAs/nvec/nvec_piRNA_cluster_TEs_intersect.txt")
```

Calculate number of unique piRNA clusters (unique genomic intervals in V1:V3)
```{r}
unique_clusters <- cluster_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNA_clusters <- nrow(unique_clusters)
```
3 out of 3 piRNA clusters overlap with some TE. 

Calculate how many unique piRNA clusters overlap each repeat type (V7)
```{r}
clusters_per_repeat <- cluster_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNA_clusters)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(clusters_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNA clusters overlapping", title = "Proportion of piRNA clusters overlapping RepeatMasker elements - Nvec") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/nvec/nvec_piRNAs_TEs_intersect.txt")
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```

724 unique piRNAs overlap with some TE. 

Calculate how many unique piRNAs overlap each repeat type (V7)
```{r}
pirna_per_repeat <- pirna_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNAs)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(pirna_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNAs overlapping", title = "Proportion of piRNAs overlapping RepeatMasker elements - Nvec") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/nvec/nvec_piRNAs_genes_intersect.txt", sep = "\t")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

180 piRNAs overlap both genes and TEs. 

To summarize: 
- 12963 piRNAs in Nvec
- 544 piRNAs overlap with some TE (724-180)
- 12783 piRNAs overlap with genes (12963-180)
- 180 piRNAs overlap with both TE and genes 
- 544+12783+180 = 13507 piRNAs overlapping TEs, genes or both 

This equals about ~44% of the piRNAs. The rest are in intergenic space (could be in other elements but these are unannotated). This matches Figure 7 ovary piRNA targets in Hydra ovaries (Praher et al. 2017).

### piRNA gene overlap enrichment 

Perform GO enrichment for genes that are overlapping with piRNAs. 

#### Make list of genes for TopGO

Read in gene to go information
```{r}
nvec_gene2go<-read.delim("../data/annotations/Nvec_gene2go.tab", sep="\t")
```

Isolate gene names in `pirna_gene`
```{r}
pirna_gene <- pirna_gene %>%
  mutate(gene_name = str_extract(V12, "(?<=ID=)[^;]+"))
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
nvec_clust_genes <- as.character(unique(pirna_gene$gene_name))

# All genes 
nvec_all_genes <- as.character(nvec_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
nvec_GeneList <- factor(as.integer(nvec_all_genes %in% nvec_clust_genes))
names(nvec_GeneList) <- nvec_all_genes
```

Read in gene-to-go-mappings
```{r}
nvec_gene2go_topgo<-readMappings("../data/annotations/Nvec_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

#### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_NVEC_BP <-new("topGOdata", ontology="BP", gene2GO=nvec_gene2go_topgo, allGenes=nvec_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_NVEC_BP_FE <- runTest(GO_NVEC_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_NVEC_BP_En <- GenTable(GO_NVEC_BP, Fisher = GO_NVEC_BP_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_NVEC_BP_En$Fisher<-as.numeric(GO_NVEC_BP_En$Fisher)
GO_NVEC_BP_En_sig<-GO_NVEC_BP_En[GO_NVEC_BP_En$Fisher<0.05,]
```

#### Molecular Functions 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_NVEC_MF <-new("topGOdata", ontology="MF", gene2GO=nvec_gene2go_topgo, allGenes=nvec_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_NVEC_MF_FE <- runTest(GO_NVEC_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_NVEC_MF_En <- GenTable(GO_NVEC_MF, Fisher = GO_NVEC_MF_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_NVEC_MF_En$Fisher<-as.numeric(GO_NVEC_MF_En$Fisher)
GO_NVEC_MF_En_sig<-GO_NVEC_MF_En[GO_NVEC_MF_En$Fisher<0.05,]
```

#### Plotting enrichment 

Plot top 10 terms
```{r}
## Bind dfs 
GO_NVEC_BP_En_sig$ontology <- "Biological Processes"
GO_NVEC_MF_En_sig$ontology <- "Molecular Functions"
BP_MF_NVEC <- rbind(GO_NVEC_BP_En_sig, GO_NVEC_MF_En_sig)

## Select top 10 terms for each ontology 
top10_per_ontology <- BP_MF_NVEC %>%
  group_by(ontology) %>%
  slice_max(Significant, n = 10, with_ties = FALSE) %>%
  ungroup()
  
ggplot(top10_per_ontology, aes(x = reorder(Term, Significant), y = Significant, fill = Fisher)) +
  geom_bar(stat = "identity") +
  #geom_text(aes(label = gene_count), vjust = -0.5, color = "black", size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  labs(x = "GO Term",
       y = "Number of Significant Genes",
       fill = "Fisher Value") +
  theme_bw()
```

## Ahya 

Read in repeatmasker out file 
```{r}
out_file <- read.table("~/Desktop/GFFs/ahya/Ahyacinthus.chrsV1.fasta.out", comment.char = "", header = F, fill = T, skip = 2)
```

Retain only chromosome, start, stop, TE name, Family Name and percent divergence
```{r}
bed_file <- out_file[,c(5:7,11,2)]

#remove empty lines 
bed_file <- bed_file %>%
  drop_na(.)
```

Calculate the proportion of each repeat type 
```{r}
bed_file <- bed_file %>%
  mutate(length = V7 - V6 + 1)  # Calculate length of each interval

# Summarize total length per repeat type
length_per_repeat <- bed_file %>%
  group_by(V11) %>%
  summarise(total_length = sum(length)) %>%
  ungroup()

# Calculate total length of all repeats
total_length_all <- sum(length_per_repeat$total_length)

# Calculate proportion of each repeat type
length_per_repeat <- length_per_repeat %>%
  mutate(proportion = total_length / total_length_all) %>%
  mutate(proportion_precent = proportion*100)

# Assume length_per_repeat contains columns: V11 (repeat type), proportion (calculated proportion)
ggplot(length_per_repeat, aes(x = reorder(V11, -proportion_precent), y = proportion_precent, fill = V11)) +
  geom_bar(stat = "identity") +                       # Bar heights represent proportions
  #scale_y_continuous(labels = scales::percent) +     # Y-axis as percentages
  labs(x = "Repeat Element Type", y = "Proportion of Genome (%)", 
       title = "Proportion of Genomic Repeat Elements") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x labels for readability
```

Write out bed file
```{r}
write.table(bed_file, file = "~/Desktop/GFFs/ahya/Ahya_repeatmasker.bed", sep = "\t", col.names = F, row.names = F, quote = F)
```

### piRNA clusters 

Read in bed overlap file
```{r}
cluster_bed <- read.table("../output/piRNAs/ahya/Ahya_piRNA_cluster_TEs_intersect.txt")
```

Calculate number of unique piRNA clusters (unique genomic intervals in V1:V3)
```{r}
unique_clusters <- cluster_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNA_clusters <- nrow(unique_clusters)
```

18 out of 29 piRNA clusters overlap with some TE. 

Calculate how many unique piRNA clusters overlap each repeat type (V7)
```{r}
clusters_per_repeat <- cluster_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNA_clusters)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(clusters_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNA clusters overlapping", title = "Proportion of piRNA clusters overlapping RepeatMasker elements - Ahya") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

Read in piRNA cluster-gene overlap 
```{r}
cluster_gene <- read.table("../output/piRNAs/ahya/Ahya_piRNA_cluster_genes_intersect.txt")
```

Calculate number of unique piRNA clusters overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_cluster_gene <- cluster_gene %>%
  distinct(V1, V2, V3)

num_unique_cluster_gene <- nrow(unique_cluster_gene)

## unique gene
unique_gene <- cluster_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/ahya/Ahya_piRNAs_TEs_intersect.txt")
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```

Calculate how many unique piRNAs overlap each repeat type (V7)
```{r}
pirna_per_repeat <- pirna_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNAs)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(pirna_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNAs overlapping", title = "Proportion of piRNAs overlapping RepeatMasker elements - Ahya") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/ahya/Ahya_piRNAs_genes_intersect.txt")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

2727 piRNAs overlap with both TEs and genes. 

To summarize: 
- 364493 piRNAs in Ahya
- 5613 piRNAs overlap with some TE
- 116936 piRNAs overlap with genes
- 2727 piRNAs overlap with both TE and genes 
- 2727+116936+5613 = 125276 piRNAs overlapping TEs, genes or both 

This equals about ~34% of the piRNAs. The rest are in intergenic space (could be in other elements but these are unannotated). This matches Figure 7 ovary piRNA targets in Hydra ovaries (Praher et al. 2017).

### piRNA gene overlap enrichment 

Perform GO enrichment for genes that are overlapping with piRNAs. 

#### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
ahya_gene2go<-read.delim("../data/annotations/Ahya_gene2go.tab", sep="\t")
```

Isolate gene names in `pirna_gene`
```{r}
pirna_gene <- pirna_gene %>%
  mutate(gene_name = str_extract(V12, "(?<=ID=)[^;]+"))
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
ahya_clust_genes <- as.character(unique(pirna_gene$gene_name))

# All genes 
ahya_all_genes <- as.character(ahya_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
ahya_GeneList <- factor(as.integer(ahya_all_genes %in% ahya_clust_genes))
names(ahya_GeneList) <- ahya_all_genes
```

Read in gene-to-go-mappings
```{r}
ahya_gene2go_topgo<-readMappings("../data/annotations/Ahya_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

#### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_AHYA_BP <-new("topGOdata", ontology="BP", gene2GO=ahya_gene2go_topgo, allGenes=ahya_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_AHYA_BP_FE <- runTest(GO_AHYA_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_AHYA_BP_En <- GenTable(GO_AHYA_BP, Fisher = GO_AHYA_BP_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_AHYA_BP_En$Fisher<-as.numeric(GO_AHYA_BP_En$Fisher)
GO_AHYA_BP_En_sig<-GO_AHYA_BP_En[GO_AHYA_BP_En$Fisher<0.05,]
```

#### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_AHYA_MF <-new("topGOdata", ontology="MF", gene2GO=ahya_gene2go_topgo, allGenes=ahya_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_AHYA_MF_FE <- runTest(GO_AHYA_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_AHYA_MF_En <- GenTable(GO_AHYA_MF, Fisher = GO_AHYA_MF_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_AHYA_MF_En$Fisher<-as.numeric(GO_AHYA_MF_En$Fisher)
GO_AHYA_MF_En_sig<-GO_AHYA_MF_En[GO_AHYA_MF_En$Fisher<0.05,]
```

#### Plotting enrichment 

Plot top 10 terms
```{r}
## Bind dfs 
GO_AHYA_BP_En_sig$ontology <- "Biological Processes"
GO_AHYA_MF_En_sig$ontology <- "Molecular Functions"
BP_MF_AHYA <- rbind(GO_AHYA_BP_En_sig, GO_AHYA_MF_En_sig)

## Select top 10 terms for each ontology 
top10_per_ontology <- BP_MF_AHYA %>%
  group_by(ontology) %>%
  slice_max(Significant, n = 10, with_ties = FALSE) %>%
  ungroup()
  
ggplot(top10_per_ontology, aes(x = reorder(Term, Significant), y = Significant, fill = Fisher)) +
  geom_bar(stat = "identity") +
  #geom_text(aes(label = gene_count), vjust = -0.5, color = "black", size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  labs(x = "GO Term",
       y = "Number of Significant Genes",
       fill = "Fisher Value") +
  theme_bw()
```

## Species comparisons

Plot percent of piRNAs overlapping with TEs, genes, both or intergenic regions 
```{r}
# Data frame with counts
percent <- data.frame(
  Species = rep(c("Acropora hyacinthus", "Astrangia poculata", "Nematostella vectensis"), each = 4),
  Element = rep(c("TEs only", "Genes only", "Both", "Intergenic"), times = 3),
  Count = c(5613, 116936, 2727, 239217, 
            59, 3037, 142, 2299,
            544, 12783, 180, 16964))

# Calculate percentages within each species
percent <- percent %>%
  group_by(Species) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

# Calculate total counts per species for labels
totals <- percent %>%
  group_by(Species) %>%
  summarize(Total = sum(Count))

# Ensure the order of factors is correct
percent$Element <- factor(percent$Element, 
                                levels = c("Intergenic", "Genes only", "Both", "TEs only"))

pirna_overlap_spp <- ggplot(percent, aes(x = Species, y = Percent, fill = Element)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", position = position_stack(reverse = T)) +
  scale_fill_brewer(palette = "Greens", direction = -1) +
  # geom_text(
  #   data = totals,
  #   aes(x = Species, y = 103, label = Total),
  #   inherit.aes = FALSE,    
  #   size = 3,
  #   color = "black") +
  scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 100, by = 10),
        limits = c(0, 105)) +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  coord_cartesian(clip = "off") +
  ylab("Percentage (%)") +
  xlab("") +
  theme_classic() + 
  theme(axis.text = element_text(color = "black", size = 20),
    axis.title = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(face = "italic"),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 15),
    legend.key.size = unit(0.6, "cm")) +
  guides(fill = guide_legend(reverse = TRUE)); pirna_overlap_spp

ggsave("../output/piRNAs/cross-species/piRNA_overlaps.pdf",
       pirna_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/cross-species/piRNA_overlaps.png",
       pirna_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
```

### Functional comparisons 

Compare terms identified in GO enrichment above across species.

Add species name in dfs 
```{r}
BP_MF_AHYA$Species <- "Acropora hyacinthus"
BP_MF_APOC$Species <- "Astrangia poculata"
BP_MF_NVEC$Species <- "Nematostella vectensis"

all_spp_go <- rbind(BP_MF_AHYA, BP_MF_APOC, BP_MF_NVEC)
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GO.IDs
shared_go_pi <- all_spp_go %>%
  group_by(GO.ID) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GO.ID)

# Create a new dataframe with only the shared GO.IDs
shared_go_pi_df <- all_spp_go %>%
  filter(GO.ID %in% shared_go_pi$GO.ID)
```

Plot! A couple of options
```{r}
GO_plot_pi1 <- ggplot(shared_go_pi_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 3, color = "black", stroke = 0.1) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  coord_flip() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw(); GO_plot_pi1 # this is what I used in my Moorea ncRNA paper 

ggplot(shared_go_pi_df, aes(x = Term, y = -log10(Fisher), fill = Species)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  ylab(expression('-log'[10]*'(Fisher p-value)')) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(shared_go_pi_df, aes(x = Term, y = -log10(Fisher), fill = Species)) +
  geom_bar(stat = "identity") +
  ylab(expression('-log'[10]*'(Fisher p-value)')) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(shared_go_pi_df, aes(x = Species, y = Term, color = -log10(Fisher), size = -log10(Fisher))) +
  geom_point() +
  scale_color_viridis_c() +
  labs(y = "Term", color = expression('(Fisher p-value)')) +
  theme_classic()

tile_go <- ggplot(shared_go_pi_df, aes(x = Species, y = Term, fill = Fisher)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "A", name = expression('Fisher p-value')) +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
        strip.text = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 11, color = "black"),
        axis.title.y = element_text(size = 13, face = "bold"),
        axis.title.x = element_text(size = 13, face = "bold")) +
  xlab("") + 
  ylab("GO Term"); tile_go

ggsave("../output/piRNAs/cross-species/piRNA_gene_overlap_enrichment.pdf",
       tile_go, width = 10, height = 9, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/cross-species/piRNA_gene_overlap_enrichment.png",
       tile_go, width = 10, height = 9, units = "in", dpi = 600, bg = "white")
```

I think I like the tile plot the best. 

Plot only terms that show up in all three species. 
```{r}
shared_terms <- shared_go_pi_df %>%
  group_by(Term) %>%
  filter(n_distinct(Species) == 3) %>%
  ungroup()

# Now plot as before
ggplot(shared_terms, aes(x = Species, y = Term, fill = -log10(Fisher))) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "D", name = expression('-log'[10]*'(Fisher p-value)')) +
  facet_wrap(~ontology, scales = "free_y", ncol = 1) +
  xlab("Species") +
  ylab("GO Term") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
        strip.text = element_text(size = 12, face = "bold"))
```



