---
title: "TE in piRNA clusters"
author: "Jill Ashey"
date: "2025-11-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(GenomicRanges)
library(topGO)
```

## Apoc 

Read in repeatmasker out file 
```{r}
out_file <- read.table("~/Desktop/GFFs/apoc/apoculata.genome.fasta.out", comment.char = "", header = F, fill = T, skip = 2)
```

Retain only chromosome, start, stop, TE name, Family Name and percent divergence
```{r}
bed_file <- out_file[,c(5:7,11,2)]

#remove empty lines 
bed_file <- bed_file %>%
  drop_na(.)
```

Calculate the proportion of each repeat type 
```{r}
bed_file <- bed_file %>%
  mutate(length = V7 - V6 + 1)  # Calculate length of each interval

# Summarize total length per repeat type
length_per_repeat <- bed_file %>%
  group_by(V11) %>%
  summarise(total_length = sum(length)) %>%
  ungroup()

# Calculate total length of all repeats
total_length_all <- sum(length_per_repeat$total_length)

# Calculate proportion of each repeat type
length_per_repeat <- length_per_repeat %>%
  mutate(proportion = total_length / total_length_all) %>%
  mutate(proportion_precent = proportion*100)

# Assume length_per_repeat contains columns: V11 (repeat type), proportion (calculated proportion)
ggplot(length_per_repeat, aes(x = reorder(V11, -proportion_precent), y = proportion_precent, fill = V11)) +
  geom_bar(stat = "identity") +                       # Bar heights represent proportions
  #scale_y_continuous(labels = scales::percent) +     # Y-axis as percentages
  labs(x = "Repeat Element Type", y = "Proportion of Genome (%)", 
       title = "Proportion of Genomic Repeat Elements") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x labels for readability
```

Write out bed file
```{r}
write.table(bed_file, file = "~/Desktop/GFFs/apoc/Apoc_repeatmasker.bed", sep = "\t", col.names = F, row.names = F, quote = F)
```

### piRNA clusters

Read in bed overlap file
```{r}
cluster_bed <- read.table("../output/piRNAs/apoc/Apoc_piRNA_cluster_TEs_intersect.txt")
```

Calculate number of unique piRNA clusters (unique genomic intervals in V1:V3)
```{r}
unique_clusters <- cluster_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNA_clusters <- nrow(unique_clusters)
```

3 piRNA clusters overlap with some TE. That is so little!!! 

Calculate how many unique piRNA clusters overlap each repeat type (V7)
```{r}
clusters_per_repeat <- cluster_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNA_clusters)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(clusters_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNA clusters overlapping", title = "Proportion of piRNA clusters overlapping RepeatMasker elements - Apoc") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

And they only overlap simple repeats. 

Read in piRNA cluster-gene overlap 
```{r}
cluster_gene <- read.table("../output/piRNAs/apoc/Apoc_piRNA_cluster_genes_intersect.txt")
```

Calculate number of unique piRNA clusters overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_cluster_gene <- cluster_gene %>%
  distinct(V1, V2, V3)

num_unique_cluster_gene <- nrow(unique_cluster_gene)

## unique gene
unique_gene <- cluster_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/apoc/Apoc_piRNAs_TEs_intersect.txt")
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```

Calculate how many unique piRNAs overlap each repeat type (V7)
```{r}
pirna_per_repeat <- pirna_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNAs)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(pirna_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNAs overlapping", title = "Proportion of piRNAs overlapping RepeatMasker elements - Apoc") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/apoc/Apoc_piRNAs_genes_intersect.txt")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

142 piRNAs overlap with both TEs and genes. 

To summarize: 
- 5537 piRNAs in Apoc
- 201 piRNAs overlap with some TE
- 3179 piRNAs overlap with genes
- 142 piRNAs overlap with both TE and genes 
- 201+3179+142 = 3522 piRNAs overlapping TEs, genes or both 

This equals about ~63% of the piRNAs. The rest are in intergenic space (could be in other elements but these are unannotated). Much less unique piRNAs in Apoc than in Ahya. 

## Nvec 

Read in repeatmasker out file
```{r}
out_file <- read.table("~/Desktop/GFFs/nvec/Nvec200.fasta.out", comment.char = "", header = F, fill = T, skip = 2)
```

Retain only chromosome, start, stop, TE name, Family Name and percent divergence
```{r}
bed_file <- out_file[,c(5:7,11,2)]

#remove empty lines 
bed_file <- bed_file %>%
  drop_na(.)
```

Write out bed file
```{r}
write.table(bed_file, file = "~/Desktop/GFFs/nvec/nvec_repeatmasker.bed", sep = "\t", col.names = F, row.names = F, quote = F)
```

### piRNA clusters 

Read in bed overlap file
```{r}
cluster_bed <- read.table("../output/piRNAs/nvec/nvec_piRNA_cluster_TEs_intersect.txt")
```

Calculate number of unique piRNA clusters (unique genomic intervals in V1:V3)
```{r}
unique_clusters <- cluster_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNA_clusters <- nrow(unique_clusters)
```
3 out of 3 piRNA clusters overlap with some TE. 

Calculate how many unique piRNA clusters overlap each repeat type (V7)
```{r}
clusters_per_repeat <- cluster_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNA_clusters)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(clusters_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNA clusters overlapping", title = "Proportion of piRNA clusters overlapping RepeatMasker elements - Nvec") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/nvec/nvec_piRNAs_TEs_intersect.txt")
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```

724 unique piRNAs overlap with some TE. 

Calculate how many unique piRNAs overlap each repeat type (V7)
```{r}
pirna_per_repeat <- pirna_bed %>%
  distinct(V1, V2, V3, V7) %>%
  group_by(V7) %>%
  summarise(num_clusters = n()) %>%
  ungroup() %>%
  mutate(proportion = num_clusters / num_unique_piRNAs)

# Plot the proportions of piRNA clusters overlapping each repeat type
ggplot(pirna_per_repeat, aes(x = reorder(V7, -proportion), y = proportion, fill = V7)) +
  geom_bar(stat = "identity") +
  labs(x = "Repeat type", y = "Proportion of piRNAs overlapping", title = "Proportion of piRNAs overlapping RepeatMasker elements - Nvec") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/nvec/nvec_piRNAs_genes_intersect.txt", sep = "\t")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

180 piRNAs overlap both genes and TEs. 

To summarize: 
- 12963 piRNAs in Nvec
- 544 piRNAs overlap with some TE (724-180)
- 12783 piRNAs overlap with genes (12963-180)
- 180 piRNAs overlap with both TE and genes 
- 544+12783+180 = 13507 piRNAs overlapping TEs, genes or both 

This equals about ~44% of the piRNAs. The rest are in intergenic space (could be in other elements but these are unannotated). This matches Figure 7 ovary piRNA targets in Hydra ovaries (Praher et al. 2017).

## Ahya 

### piRNAs

Read in piRNA TE overlap 
```{r}
pirna_bed <- read.table("../output/piRNAs/ahya/Ahya_piRNAs_TEs_repeats_intersect.txt")
```

Calculate number of unique piRNAs (unique genomic intervals in V1:V3)
```{r}
unique_pirna <- pirna_bed %>%
  distinct(V1, V2, V3)

num_unique_piRNAs <- nrow(unique_pirna)
```

Look at repeats and piRNA overlap 
```{r}
pirna_summary <- pirna_bed %>%
  mutate(piRNA_ID = paste(V1, V2, V3, sep = "_")) %>%
  mutate(Repeat_Family = str_extract(V12, "(?<=genus:)[^|;]+")) %>%
  distinct(piRNA_ID, Repeat_Family, .keep_all = TRUE) %>%
  mutate(Repeat_Group = substr(Repeat_Family, 1, 4))

total_unique_pirnas <- n_distinct(pirna_summary$piRNA_ID)
cat("Total unique piRNAs overlapping repeats:", total_unique_pirnas, "\n")

# Count piRNAs per repeat type/family
repeat_counts <- pirna_summary %>%
  count(Repeat_Group, sort = TRUE, name = "piRNA_Count") %>%
  mutate(Species = "Acropora hyacinthus")
```

Summarize repeat data further 
```{r}
# Clean up messy names 
repeat_clean <- repeat_counts %>%
  mutate(
    Clean_Group = case_when(
      startsWith(Repeat_Group, "DNA") ~ "DNA Transposon",
      startsWith(Repeat_Group, "LTR") ~ "LTR Retrotransposon",
      startsWith(Repeat_Group, "LINE") ~ "LINE",
      startsWith(Repeat_Group, "Unkn") ~ "Unknown",
      startsWith(Repeat_Group, "Simp") ~ "Simple Repeat",
      startsWith(Repeat_Group, "Low")  ~ "Low Complexity",
      startsWith(Repeat_Group, "RC")   ~ "Rolling Circle (Helitron)",
      TRUE ~ "Other")) %>%
  group_by(Species, Clean_Group) %>%
  summarise(piRNA_Count = sum(piRNA_Count), .groups = "drop")

# Plot 
ggplot(repeat_clean, aes(x = Species, y = piRNA_Count, fill = Clean_Group)) +
  geom_col(position = "fill", width = 0.5, color = "black", linewidth = 0.2) +
  #scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = NULL,
    y = "Percentage of Total piRNAs",
    fill = "Repeat Class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(face = "italic"),
    panel.grid.major.x = element_blank())
```

Read in piRNA-gene overlap 
```{r}
pirna_gene <- read.table("../output/piRNAs/ahya/Ahya_piRNAs_genes_intersect.txt")
```

Calculate lengths
```{r}
pirna_gene <- pirna_gene %>%
  mutate(pirna_length = V3 - V2) %>%
  mutate(gene_length = V8 - V7)
```

Calculate number of unique piRNAs overlapping genes (unique genomic intervals in V1:V3) and vice versa (unique genomic intervals in V4, V6, V7)
```{r}
## unique pirna
unique_pirna_gene <- pirna_gene %>%
  distinct(V1, V2, V3)

num_unique_piRNAs_gene <- nrow(unique_pirna_gene)

## unique gene
unique_gene <- pirna_gene %>%
  distinct(V4, V7, V8)

num_unique_gene <- nrow(unique_gene)
```

Is there overlap between the piRNAs overlapping genes v. TEs?
```{r}
# Create GRanges objects
gr_gene <- GRanges(seqnames = unique_pirna_gene$V1,
                   ranges = IRanges(start = unique_pirna_gene$V2, end = unique_pirna_gene$V3))

gr_te <- GRanges(seqnames = unique_pirna$V1,
                 ranges = IRanges(start = unique_pirna$V2, end = unique_pirna$V3))

# Find overlaps
overlaps <- findOverlaps(gr_gene, gr_te)

# Get the overlapping intervals (from each data frame)
overlapping_genes <- unique_pirna_gene[queryHits(overlaps), ]
overlapping_tes <- unique_pirna[subjectHits(overlaps), ]

# How many overlaps found?
length(overlaps)
```

72962 piRNAs overlap with both TEs and genes. 

To summarize: 
- 364493 piRNAs in Ahya
- 265966 piRNAs overlap with some TE
- 119663 piRNAs overlap with genes
- 72962 piRNAs overlap with both TE and genes 

## Species comparisons

Plot percent of piRNAs overlapping with TEs, genes, both or intergenic regions 
```{r}
# Data frame with counts
percent <- data.frame(
  Species = rep(c("Acropora hyacinthus", "Astrangia poculata", "Nematostella vectensis"), each = 4),
  Element = rep(c("TEs only", "Genes only", "Both", "Intergenic"), times = 3),
  Count = c(193004, 46701, 72962, 51826, 
            59, 3037, 142, 2299,
            544, 12783, 180, 16964))

# Calculate percentages within each species
percent <- percent %>%
  group_by(Species) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

# Calculate total counts per species for labels
totals <- percent %>%
  group_by(Species) %>%
  summarize(Total = sum(Count))

# Ensure the order of factors is correct
percent$Element <- factor(percent$Element, 
                                levels = c("Intergenic", "Genes only", "Both", "TEs only"))

pirna_overlap_spp <- ggplot(percent, aes(x = Species, y = Percent, fill = Element)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", position = position_stack(reverse = T)) +
  scale_fill_brewer(palette = "Greens", direction = -1) +
  # geom_text(
  #   data = totals,
  #   aes(x = Species, y = 103, label = Total),
  #   inherit.aes = FALSE,    
  #   size = 3,
  #   color = "black") +
  scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 100, by = 10),
        limits = c(0, 105)) +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  coord_cartesian(clip = "off") +
  ylab("Percentage (%)") +
  xlab("") +
  theme_classic() + 
  theme(axis.text = element_text(color = "black", size = 20),
    axis.title = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(face = "italic"),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 15),
    legend.key.size = unit(0.6, "cm")) +
  guides(fill = guide_legend(reverse = TRUE)); pirna_overlap_spp

ggsave("../output/piRNAs/cross-species/piRNA_overlaps.pdf",
       pirna_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
ggsave("../output/piRNAs/cross-species/piRNA_overlaps.png",
       pirna_overlap_spp, width = 10, height = 6, units = "in", dpi = 600, bg = "white")
```






