---
title: "Homology of smRNA targets"
author: "Jill Ashey"
date: "2025-11-07"
output: html_document
---

This script will examine the homology between the miRNA targets and the genes overlapping piRNAs. Protein homology was done via broccoli. Using Zoe Dellaert's [code](https://github.com/zdellaert/multi-sp-snRNA/blob/main/reference_genes/Marker-Genes.md) for reference. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stringr)
library(janitor)
require("rtracklayer")
library(UpSetR)
library(ggupset)
library(data.table)
library(patchwork)
```

## miRNA targets 

Read in data for all spp - mRNA only here
```{r}
## Ahya
miranda_ahya <- read.delim("../output/miRNA_interactions/ahya/miranda_strict_mRNA_parsed_ahya.txt", header = F)
colnames(miranda_ahya) <- c("miRNA", "mRNA", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

## Apoc 
miranda_apoc <- read.delim("../output/miRNA_interactions/apoc/miranda_strict_mRNA_parsed_apoc.txt", header = F)
colnames(miranda_apoc) <- c("miRNA", "mRNA", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")

## Nvec 
miranda_nvec <- read.delim("../output/miRNA_interactions/nvec/miranda_strict_mRNA_parsed_nvec.txt", header = F)
colnames(miranda_nvec) <- c("miRNA", "mRNA", "score", "energy", "query_start_end", "subject_start_end", "total_bp_shared", "query_similar", "subject_similar")
```

Subset mRNA names
```{r}
mRNA_targets_ahya <- unique(miranda_ahya$mRNA)
mRNA_targets_apoc <- unique(miranda_apoc$mRNA)
mRNA_targets_nvec <- unique(miranda_nvec$mRNA)
```

Read in broccoli results 
```{r}
broc_groups <- data.table::fread("../output/broccoli/dir_step3/table_OGs_protein_names.txt", header = T)
```

Identify targets in broccoli data 
```{r}
# Function to extract gene IDs from protein IDs
extract_gene_id <- function(protein_ids, species) {
  if (species == "nvec") {
    # Remove "protein|" prefix if present, keep the base ID
    gsub("protein\\|", "", protein_ids)
  } else if (species == "ahya") {
    # Keep as is
    protein_ids
  } else if (species == "apoc") {
    # Remove "protein|" prefix if present
    gsub("protein\\|", "", protein_ids)
  }
}

# Extract gene IDs from the broc_groups columns
broc_groups[, Apoc_genes := lapply(strsplit(Apoc_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "apoc"))]
broc_groups[, Nvec_genes := lapply(strsplit(Nvec_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "nvec"))]
broc_groups[, Ahya_genes := lapply(strsplit(Ahya_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "ahya"))]

# Find which OGs contain targets from each species
broc_groups[, has_nvec_target := sapply(Nvec_genes, function(genes) 
  any(mRNA_targets_nvec %in% genes))]
broc_groups[, has_ahya_target := sapply(Ahya_genes, function(genes) 
  any(mRNA_targets_ahya %in% genes))]
broc_groups[, has_apoc_target := sapply(Apoc_genes, function(genes) 
  any(mRNA_targets_apoc %in% genes))]

# Find OGs with targets from all three species (homologous targets)
homologous_targets <- broc_groups[has_nvec_target == TRUE & 
                                   has_ahya_target == TRUE & 
                                   has_apoc_target == TRUE]

# Find OGs with targets from any two species
two_species <- broc_groups[(has_nvec_target + has_ahya_target + has_apoc_target) == 2]

# Find OGs with targets from only one species
one_species <- broc_groups[(has_nvec_target + has_ahya_target + has_apoc_target) == 1]

# Summary statistics
cat("Summary of mRNA target representation in BROCCOLI groups:\n")
cat("=========================================================\n")
cat("Total OGs with homologous targets (all 3 species):", nrow(homologous_targets), "\n")
cat("Total OGs with targets from 2 species:", nrow(two_species), "\n")
cat("Total OGs with targets from 1 species:", nrow(one_species), "\n\n")

cat("Targets found per species:\n")
cat("Nvec targets in OGs:", sum(broc_groups$has_nvec_target), "/", length(mRNA_targets_nvec), "\n")
cat("Ahya targets in OGs:", sum(broc_groups$has_ahya_target), "/", length(mRNA_targets_ahya), "\n")
cat("Apoc targets in OGs:", sum(broc_groups$has_apoc_target), "/", length(mRNA_targets_apoc), "\n")

# View the homologous targets
print(homologous_targets[, .(`#OG_name`, Apoc_proteins.faa, Nvec_proteins.faa, Ahya_proteins.faa)])

# Get specific target IDs in homologous groups
homologous_targets[, nvec_target_ids := sapply(Nvec_genes, function(genes) 
  paste(genes[genes %in% mRNA_targets_nvec], collapse = ", "))]
homologous_targets[, ahya_target_ids := sapply(Ahya_genes, function(genes) 
  paste(genes[genes %in% mRNA_targets_ahya], collapse = ", "))]
homologous_targets[, apoc_target_ids := sapply(Apoc_genes, function(genes) 
  paste(genes[genes %in% mRNA_targets_apoc], collapse = ", "))]
```

Make upset plot of orthogroups with targets  
```{r}
# Function to check if a target is in any homologous group
find_homologous_groups <- function(target, species, broc_groups) {
  if (species == "nvec") {
    ogs <- broc_groups[sapply(Nvec_genes, function(genes) target %in% genes), `#OG_name`]
  } else if (species == "ahya") {
    ogs <- broc_groups[sapply(Ahya_genes, function(genes) target %in% genes), `#OG_name`]
  } else if (species == "apoc") {
    ogs <- broc_groups[sapply(Apoc_genes, function(genes) target %in% genes), `#OG_name`]
  }
  return(ogs)
}

# Create a binary matrix showing which species have targets in each OG
# Get all OGs that contain at least one target from any species
ogs_with_targets <- unique(broc_groups[has_nvec_target == TRUE | 
                                        has_ahya_target == TRUE | 
                                        has_apoc_target == TRUE, `#OG_name`])

# Create a list for UpSetR (binary membership)
upset_data <- data.frame(
  OG = ogs_with_targets,
  Nvec = 0,
  Ahya = 0,
  Apoc = 0
)

# Fill in the binary matrix
for (i in 1:nrow(upset_data)) {
  og <- upset_data$OG[i]
  upset_data$Nvec[i] <- as.integer(broc_groups[`#OG_name` == og, has_nvec_target])
  upset_data$Ahya[i] <- as.integer(broc_groups[`#OG_name` == og, has_ahya_target])
  upset_data$Apoc[i] <- as.integer(broc_groups[`#OG_name` == og, has_apoc_target])
}

# Create the UpSet plot
upset(upset_data, 
      sets = c("Nvec", "Ahya", "Apoc"),
      order.by = "freq",
      main.bar.color = "steelblue",
      sets.bar.color = c("coral", "lightblue", "lightgreen"),
      text.scale = c(1.5, 1.3, 1.2, 1, 1.5, 1.2),
      point.size = 3.5,
      line.size = 1)

# Get summary statistics for the plot
cat("\nUpSet Plot Summary:\n")
cat("===================\n")
cat("Total OGs with targets:", nrow(upset_data), "\n")
cat("OGs with all 3 species:", sum(upset_data$Nvec == 1 & upset_data$Ahya == 1 & upset_data$Apoc == 1), "\n")
cat("OGs with Nvec + Ahya:", sum(upset_data$Nvec == 1 & upset_data$Ahya == 1 & upset_data$Apoc == 0), "\n")
cat("OGs with Nvec + Apoc:", sum(upset_data$Nvec == 1 & upset_data$Ahya == 0 & upset_data$Apoc == 1), "\n")
cat("OGs with Ahya + Apoc:", sum(upset_data$Nvec == 0 & upset_data$Ahya == 1 & upset_data$Apoc == 1), "\n")
cat("OGs with Nvec only:", sum(upset_data$Nvec == 1 & upset_data$Ahya == 0 & upset_data$Apoc == 0), "\n")
cat("OGs with Ahya only:", sum(upset_data$Nvec == 0 & upset_data$Ahya == 1 & upset_data$Apoc == 0), "\n")
cat("OGs with Apoc only:", sum(upset_data$Nvec == 0 & upset_data$Ahya == 0 & upset_data$Apoc == 1), "\n")

# Optional: Create a different version with ggplot2
upset_list <- list(
  Nvec = broc_groups[has_nvec_target == TRUE, `#OG_name`],
  Ahya = broc_groups[has_ahya_target == TRUE, `#OG_name`],
  Apoc = broc_groups[has_apoc_target == TRUE, `#OG_name`]
)

# Create a tibble for ggupset
upset_tidy <- tibble(
  OG = ogs_with_targets,
  Species = lapply(ogs_with_targets, function(og) {
    species <- c()
    if (og %in% upset_list$Nvec) species <- c(species, "Nvec")
    if (og %in% upset_list$Ahya) species <- c(species, "Ahya")
    if (og %in% upset_list$Apoc) species <- c(species, "Apoc")
    return(species)
  })
)

# Create ggupset plot
ggplot(upset_tidy, aes(x = Species)) +
  geom_bar(fill = "steelblue") +
  scale_x_upset() +
  theme_minimal() +
  labs(title = "Homologous mRNA Targets Across Species",
       y = "Number of Orthology Groups") +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

Make upset plot with homologous targets 
```{r}
# For each target, identify which species have homologs in the same OG

# Initialize a list to store target information
target_membership <- list()

# Process Nvec targets
for (target in mRNA_targets_nvec) {
  # Find which OGs contain this target
  ogs <- broc_groups[sapply(Nvec_genes, function(genes) target %in% genes), `#OG_name`]
  
  if (length(ogs) > 0) {
    # For each OG this target is in, check which other species have targets
    for (og in ogs) {
      has_ahya <- broc_groups[`#OG_name` == og, has_ahya_target]
      has_apoc <- broc_groups[`#OG_name` == og, has_apoc_target]
      
      target_membership[[target]] <- list(
        target = target,
        species = "Nvec",
        OG = og,
        Nvec = 1,
        Ahya = as.integer(has_ahya),
        Apoc = as.integer(has_apoc)
      )
    }
  }
}

# Process Ahya targets
for (target in mRNA_targets_ahya) {
  ogs <- broc_groups[sapply(Ahya_genes, function(genes) target %in% genes), `#OG_name`]
  
  if (length(ogs) > 0) {
    for (og in ogs) {
      has_nvec <- broc_groups[`#OG_name` == og, has_nvec_target]
      has_apoc <- broc_groups[`#OG_name` == og, has_apoc_target]
      
      target_membership[[target]] <- list(
        target = target,
        species = "Ahya",
        OG = og,
        Nvec = as.integer(has_nvec),
        Ahya = 1,
        Apoc = as.integer(has_apoc)
      )
    }
  }
}

# Process Apoc targets
for (target in mRNA_targets_apoc) {
  ogs <- broc_groups[sapply(Apoc_genes, function(genes) target %in% genes), `#OG_name`]
  
  if (length(ogs) > 0) {
    for (og in ogs) {
      has_nvec <- broc_groups[`#OG_name` == og, has_nvec_target]
      has_ahya <- broc_groups[`#OG_name` == og, has_ahya_target]
      
      target_membership[[target]] <- list(
        target = target,
        species = "Apoc",
        OG = og,
        Nvec = as.integer(has_nvec),
        Ahya = as.integer(has_ahya),
        Apoc = 1
      )
    }
  }
}

# Convert to data frame
upset_targets <- do.call(rbind, lapply(target_membership, as.data.frame))

# Remove duplicates (targets might be in multiple OGs)
upset_targets <- upset_targets[!duplicated(upset_targets$target), ]

# Create the UpSet plot with target counts
upset(upset_targets, 
      sets = c("Nvec", "Ahya", "Apoc"),
      order.by = "freq",
      main.bar.color = "steelblue",
      sets.bar.color = c("coral", "lightblue", "lightgreen"),
      text.scale = c(1.5, 1.3, 1.2, 1, 1.5, 1.2),
      point.size = 3.5,
      line.size = 1,
      mainbar.y.label = "Number of mRNA Targets",
      sets.x.label = "Total Targets per Species")

# Print detailed statistics
cat("\nTarget-Level UpSet Summary:\n")
cat("============================\n")
cat("Total unique targets in OGs:", nrow(upset_targets), "\n\n")

cat("Targets with homologs in all 3 species:", 
    sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 1 & upset_targets$Apoc == 1), "\n")

cat("Targets with homologs in Nvec + Ahya only:", 
    sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 1 & upset_targets$Apoc == 0), "\n")

cat("Targets with homologs in Nvec + Apoc only:", 
    sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 0 & upset_targets$Apoc == 1), "\n")

cat("Targets with homologs in Ahya + Apoc only:", 
    sum(upset_targets$Nvec == 0 & upset_targets$Ahya == 1 & upset_targets$Apoc == 1), "\n")

cat("\nSpecies-specific targets (no homologs in other species):\n")
cat("Nvec only:", sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 0 & upset_targets$Apoc == 0), "\n")
cat("Ahya only:", sum(upset_targets$Nvec == 0 & upset_targets$Ahya == 1 & upset_targets$Apoc == 0), "\n")
cat("Apoc only:", sum(upset_targets$Nvec == 0 & upset_targets$Ahya == 0 & upset_targets$Apoc == 1), "\n")

# Show breakdown by species
cat("\n\nBreakdown by species:\n")
cat("Nvec targets with homologs:", sum(upset_targets$species == "Nvec"), 
    "out of", length(mRNA_targets_nvec), 
    sprintf("(%.1f%%)\n", 100 * sum(upset_targets$species == "Nvec") / length(mRNA_targets_nvec)))

cat("Ahya targets with homologs:", sum(upset_targets$species == "Ahya"), 
    "out of", length(mRNA_targets_ahya),
    sprintf("(%.1f%%)\n", 100 * sum(upset_targets$species == "Ahya") / length(mRNA_targets_ahya)))

cat("Apoc targets with homologs:", sum(upset_targets$species == "Apoc"), 
    "out of", length(mRNA_targets_apoc),
    sprintf("(%.1f%%)\n", 100 * sum(upset_targets$species == "Apoc") / length(mRNA_targets_apoc)))

# Optional: Save the target membership data for further analysis
write.csv(upset_targets, "target_homology_membership.csv", row.names = FALSE)

# Alternative visualization using ggupset
library(ggplot2)
library(ggupset)
library(tidyr)

upset_for_gg <- upset_targets %>%
  mutate(Species_set = pmap(list(Nvec, Ahya, Apoc), function(n, a, p) {
    spp <- c()
    if (n == 1) spp <- c(spp, "Nvec")
    if (a == 1) spp <- c(spp, "Ahya")
    if (p == 1) spp <- c(spp, "Apoc")
    return(list(spp))
  })) %>%
  unnest(Species_set)

ggplot(upset_for_gg, aes(x = Species_set)) +
  geom_bar(fill = "steelblue", alpha = 0.8) +
  scale_x_upset(n_intersections = 20) +
  theme_minimal() +
  labs(title = "mRNA Target Homology Across Species",
       subtitle = "Number of individual targets with homologs in each species combination",
       y = "Number of mRNA Targets") +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10))
```

### Annotation of shared/divergent targets 

Read in annotation files 
```{r}
## Apoc
annot_apoc <- read.delim("../data/annotations/apoculata_GeneAnnotation_combined.txt")

## Nvec 
annot_nvec <- read.delim("../data/annotations/NV2g.20240221.GO.tsv", header = F)
colnames(annot_nvec) <- c("gene_name", "ontology", "GeneOntologyIDs", "GO_description", "Extended_description", "Db")

## Ahya
annot_ahya <- read.delim("~/Desktop/GFFs/ahya/Acropora_hyacinthus.annotations.txt")[,1:22]
```

Merge annotation and homology info
```{r}
# Convert annotation tables to data.tables if not already
setDT(upset_targets)
upset_targets$target <- gsub("-RA", "", upset_targets$target)
setDT(annot_apoc)
setDT(annot_nvec)
setDT(annot_ahya)

# Function to merge upset_targets by species and retrieve annotations
get_annotations <- function(upset_targets, annot_apoc, annot_nvec, annot_ahya) {
  # Merge for Apoc
  apoc_annots <- merge(
    upset_targets[species == "Apoc"], 
    annot_apoc, 
    by.x = "target", 
    by.y = "Protein_ID", 
    all.x = TRUE
  )
  
  # Merge for Nvec
  nvec_annots <- merge(
    upset_targets[species == "Nvec"], 
    annot_nvec, 
    by.x = "target", 
    by.y = "gene_name", 
    all.x = TRUE
  )
  
  # Merge for Ahya
  ahya_annots <- merge(
    upset_targets[species == "Ahya"], 
    annot_ahya, 
    by.x = "target", 
    by.y = "GeneID", 
    all.x = TRUE
  )
  
  # Combine all results
  rbindlist(list(apoc_annots, nvec_annots, ahya_annots), fill = TRUE)
}

# Get a combined table of upset targets and their annotations
annotated_upset_targets <- get_annotations(upset_targets, annot_apoc, annot_nvec, annot_ahya)

# View results
head(annotated_upset_targets)
```

## piRNAs overlapping with genes 

Read in data for all spp
```{r}
## Ahya
pirna_gene_ahya <- read.table("../output/piRNAs/ahya/Ahya_piRNAs_genes_intersect.txt")

## Apoc 
pirna_gene_apoc <- read.table("../output/piRNAs/apoc/Apoc_piRNAs_genes_intersect.txt")

## Nvec
pirna_gene_nvec <- read.table("../output/piRNAs/nvec/nvec_piRNAs_genes_intersect.txt", sep = "\t")
```

Isolate gene names 
```{r}
## Ahya 
pirna_gene_ahya <- pirna_gene_ahya %>%
  mutate(gene_name = str_extract(V12, "(?<=ID=)[^;]+"))
pirna_gene_ahya$gene_name <- paste0(pirna_gene_ahya$gene_name, "-RA")

## Apoc 
pirna_gene_apoc <- pirna_gene_apoc %>%
  mutate(gene_name = str_extract(V12, "(?<=ID=)[^;]+"))
pirna_gene_apoc$gene_name <- gsub("TU", "model", pirna_gene_apoc$gene_name)

## Nvec 
pirna_gene_nvec <- pirna_gene_nvec %>%
  mutate(gene_name = str_extract(V12, "(?<=ID=)[^;]+"))
```

Subset mRNA/gene names 
```{r}
mRNA_overlap_ahya <- unique(pirna_gene_ahya$gene_name)
mRNA_overlap_apoc <- unique(pirna_gene_apoc$gene_name)
mRNA_overlap_nvec <- unique(pirna_gene_nvec$gene_name)
```

Read in broccoli results 
```{r}
broc_groups <- data.table::fread("../output/broccoli/dir_step3/table_OGs_protein_names.txt", header = T)
```

Identify targets in broccoli data 
```{r}
# Function to extract gene IDs from protein IDs
extract_gene_id <- function(protein_ids, species) {
  if (species == "nvec") {
    # Remove "protein|" prefix if present, keep the base ID
    gsub("protein\\|", "", protein_ids)
  } else if (species == "ahya") {
    # Keep as is
    protein_ids
  } else if (species == "apoc") {
    # Remove "protein|" prefix if present
    gsub("protein\\|", "", protein_ids)
  }
}

# Extract gene IDs from the broc_groups columns
broc_groups[, Apoc_genes := lapply(strsplit(Apoc_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "apoc"))]
broc_groups[, Nvec_genes := lapply(strsplit(Nvec_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "nvec"))]
broc_groups[, Ahya_genes := lapply(strsplit(Ahya_proteins.faa, " "), 
                                     function(x) extract_gene_id(x, "ahya"))]

# Find which OGs contain targets from each species
broc_groups[, has_nvec_target := sapply(Nvec_genes, function(genes) 
  any(mRNA_overlap_nvec %in% genes))]
broc_groups[, has_ahya_target := sapply(Ahya_genes, function(genes) 
  any(mRNA_overlap_ahya %in% genes))]
broc_groups[, has_apoc_target := sapply(Apoc_genes, function(genes) 
  any(mRNA_overlap_apoc %in% genes))]

# Find OGs with targets from all three species (homologous targets)
homologous_targets <- broc_groups[has_nvec_target == TRUE & 
                                   has_ahya_target == TRUE & 
                                   has_apoc_target == TRUE]

# Find OGs with targets from any two species
two_species <- broc_groups[(has_nvec_target + has_ahya_target + has_apoc_target) == 2]

# Find OGs with targets from only one species
one_species <- broc_groups[(has_nvec_target + has_ahya_target + has_apoc_target) == 1]

# Summary statistics
cat("Summary of mRNA target representation in BROCCOLI groups:\n")
cat("=========================================================\n")
cat("Total OGs with homologous targets (all 3 species):", nrow(homologous_targets), "\n")
cat("Total OGs with targets from 2 species:", nrow(two_species), "\n")
cat("Total OGs with targets from 1 species:", nrow(one_species), "\n\n")

cat("Targets found per species:\n")
cat("Nvec targets in OGs:", sum(broc_groups$has_nvec_target), "/", length(mRNA_overlap_nvec), "\n")
cat("Ahya targets in OGs:", sum(broc_groups$has_ahya_target), "/", length(mRNA_overlap_ahya), "\n")
cat("Apoc targets in OGs:", sum(broc_groups$has_apoc_target), "/", length(mRNA_overlap_apoc), "\n")

# View the homologous targets
print(homologous_targets[, .(`#OG_name`, Apoc_proteins.faa, Nvec_proteins.faa, Ahya_proteins.faa)])

# Get specific target IDs in homologous groups
homologous_targets[, nvec_target_ids := sapply(Nvec_genes, function(genes) 
  paste(genes[genes %in% mRNA_overlap_nvec], collapse = ", "))]
homologous_targets[, ahya_target_ids := sapply(Ahya_genes, function(genes) 
  paste(genes[genes %in% mRNA_overlap_ahya], collapse = ", "))]
homologous_targets[, apoc_target_ids := sapply(Apoc_genes, function(genes) 
  paste(genes[genes %in% mRNA_overlap_apoc], collapse = ", "))]
```

Make upset plot with homologous genes overlapping with piRNAs 
```{r}
# For each target, identify which species have homologs in the same OG

# Initialize a list to store target information
target_membership <- list()

# Process Nvec targets
for (target in mRNA_overlap_nvec) {
  # Find which OGs contain this target
  ogs <- broc_groups[sapply(Nvec_genes, function(genes) target %in% genes), `#OG_name`]
  
  if (length(ogs) > 0) {
    # For each OG this target is in, check which other species have targets
    for (og in ogs) {
      has_ahya <- broc_groups[`#OG_name` == og, has_ahya_target]
      has_apoc <- broc_groups[`#OG_name` == og, has_apoc_target]
      
      target_membership[[target]] <- list(
        target = target,
        species = "Nvec",
        OG = og,
        Nvec = 1,
        Ahya = as.integer(has_ahya),
        Apoc = as.integer(has_apoc)
      )
    }
  }
}

# Process Ahya targets
for (target in mRNA_overlap_ahya) {
  ogs <- broc_groups[sapply(Ahya_genes, function(genes) target %in% genes), `#OG_name`]
  
  if (length(ogs) > 0) {
    for (og in ogs) {
      has_nvec <- broc_groups[`#OG_name` == og, has_nvec_target]
      has_apoc <- broc_groups[`#OG_name` == og, has_apoc_target]
      
      target_membership[[target]] <- list(
        target = target,
        species = "Ahya",
        OG = og,
        Nvec = as.integer(has_nvec),
        Ahya = 1,
        Apoc = as.integer(has_apoc)
      )
    }
  }
}

# Process Apoc targets
for (target in mRNA_overlap_apoc) {
  ogs <- broc_groups[sapply(Apoc_genes, function(genes) target %in% genes), `#OG_name`]
  
  if (length(ogs) > 0) {
    for (og in ogs) {
      has_nvec <- broc_groups[`#OG_name` == og, has_nvec_target]
      has_ahya <- broc_groups[`#OG_name` == og, has_ahya_target]
      
      target_membership[[target]] <- list(
        target = target,
        species = "Apoc",
        OG = og,
        Nvec = as.integer(has_nvec),
        Ahya = as.integer(has_ahya),
        Apoc = 1
      )
    }
  }
}

# Convert to data frame
upset_targets <- do.call(rbind, lapply(target_membership, as.data.frame))

# Remove duplicates (targets might be in multiple OGs)
upset_targets <- upset_targets[!duplicated(upset_targets$target), ]

# Create the UpSet plot with target counts
upset(upset_targets, 
      sets = c("Nvec", "Ahya", "Apoc"),
      order.by = "freq",
      main.bar.color = "steelblue",
      #sets.bar.color = c("coral", "lightblue", "lightgreen"),
      text.scale = c(1.5, 1.3, 1.2, 1, 1.5, 1.2),
      point.size = 3.5,
      line.size = 1,
      mainbar.y.label = "Intersection Size",
      sets.x.label = "Set Size")

# Print detailed statistics
cat("\nTarget-Level UpSet Summary:\n")
cat("============================\n")
cat("Total unique targets in OGs:", nrow(upset_targets), "\n\n")

cat("Targets with homologs in all 3 species:", 
    sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 1 & upset_targets$Apoc == 1), "\n")

cat("Targets with homologs in Nvec + Ahya only:", 
    sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 1 & upset_targets$Apoc == 0), "\n")

cat("Targets with homologs in Nvec + Apoc only:", 
    sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 0 & upset_targets$Apoc == 1), "\n")

cat("Targets with homologs in Ahya + Apoc only:", 
    sum(upset_targets$Nvec == 0 & upset_targets$Ahya == 1 & upset_targets$Apoc == 1), "\n")

cat("\nSpecies-specific targets (no homologs in other species):\n")
cat("Nvec only:", sum(upset_targets$Nvec == 1 & upset_targets$Ahya == 0 & upset_targets$Apoc == 0), "\n")
cat("Ahya only:", sum(upset_targets$Nvec == 0 & upset_targets$Ahya == 1 & upset_targets$Apoc == 0), "\n")
cat("Apoc only:", sum(upset_targets$Nvec == 0 & upset_targets$Ahya == 0 & upset_targets$Apoc == 1), "\n")

# Show breakdown by species
cat("\n\nBreakdown by species:\n")
cat("Nvec targets with homologs:", sum(upset_targets$species == "Nvec"), 
    "out of", length(mRNA_overlap_nvec), 
    sprintf("(%.1f%%)\n", 100 * sum(upset_targets$species == "Nvec") / length(mRNA_overlap_nvec)))

cat("Ahya targets with homologs:", sum(upset_targets$species == "Ahya"), 
    "out of", length(mRNA_overlap_ahya),
    sprintf("(%.1f%%)\n", 100 * sum(upset_targets$species == "Ahya") / length(mRNA_overlap_ahya)))

cat("Apoc targets with homologs:", sum(upset_targets$species == "Apoc"), 
    "out of", length(mRNA_overlap_apoc),
    sprintf("(%.1f%%)\n", 100 * sum(upset_targets$species == "Apoc") / length(mRNA_overlap_apoc)))

# Optional: Save the target membership data for further analysis
#write.csv(upset_targets, "target_homology_membership.csv", row.names = FALSE)
```

Calculate and plot the percentage of each species in the shared orthogroups. 
```{r}
# Create intersection labels (binary patterns as strings for matching UpSet ordering)
upset_targets <- upset_targets %>%
  mutate(intersection = paste(Nvec, Ahya, Apoc, sep = "_"))

# Summarize counts by intersection and species
intersection_summary <- upset_targets %>%
  group_by(intersection, species) %>%
  summarise(
    count = n(),
    total = sum(Nvec + Ahya + Apoc),  # Optional: verify set membership
    .groups = "drop"
  ) %>%
  group_by(intersection) %>%
  mutate(
    pct = count / sum(count) * 100  # Percent within each intersection
  ) %>%
  ungroup()

# Create your UpSet plot (assign to object to access internals if needed)
up <- upset(
  upset_targets, 
  sets = c("Nvec", "Ahya", "Apoc"),
  #order.by = "freq",
  main.bar.color = "steelblue",
  text.scale = c(1.5, 1.3, 1.2, 1, 1.5, 1.2),
  point.size = 3.5,
  line.size = 1,
  mainbar.y.label = "Intersection Size",
  sets.x.label = "Set Size"); up

# Extract ordered intersection labels (binary patterns from UpSet matrix)
# UpSetR stores this internally; access via fromExpression or manual calculation
ordered_intersections <- upset_targets %>%
  group_by(intersection) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%  # Matches "freq" order
  pull(intersection)

# Filter summary to only shown intersections (if UpSet limits them)
intersection_summary_ordered <- intersection_summary %>%
  filter(intersection %in% ordered_intersections) %>%
  mutate(intersection = factor(intersection, levels = ordered_intersections))

species_stacked <- ggplot(intersection_summary_ordered, aes(x = intersection, y = pct, fill = species)) +
  geom_col(width = 0.8, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = c("Apoc" = "seagreen3", "Ahya" = "dodgerblue", "Nvec" = "tomato")) +  # Match your colors
  labs(
    x = "Intersections (ordered by size)",
    y = "Species % within Intersection",
    fill = "Species",
    title = "Species Composition by Intersection"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  coord_flip(); species_stacked
```

## Functional enrichment of shared genes overlapping with piRNAs 

### Apoc - shared between all three spp 

##### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
apoc_gene2go<-read.delim("../data/annotations/Apoc_gene2go.tab", sep="\t")
```

Filter by shared orthogroups and species
```{r}
upset_targets_apoc <- upset_targets %>%
  filter(
    species == "Apoc",
    Nvec == 1,
    Ahya == 1,
    Apoc == 1)
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
apoc_clust_genes <- as.character(unique(upset_targets_apoc$target))

# All genes 
apoc_all_genes <- as.character(apoc_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
apoc_GeneList <- factor(as.integer(apoc_all_genes %in% apoc_clust_genes))
names(apoc_GeneList) <- apoc_all_genes
```

Read in gene-to-go-mappings
```{r}
apoc_gene2go_topgo<-readMappings("../data/annotations/Apoc_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

##### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APOC_BP_all_spp <-new("topGOdata", ontology="BP", gene2GO=apoc_gene2go_topgo, allGenes=apoc_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APOC_BP_all_spp_FE <- runTest(GO_APOC_BP_all_spp, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APOC_BP_all_spp_En <- GenTable(GO_APOC_BP_all_spp, Fisher = GO_APOC_BP_all_spp_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APOC_BP_all_spp_En$Fisher<-as.numeric(GO_APOC_BP_all_spp_En$Fisher)
GO_APOC_BP_all_spp_En_sig<-GO_APOC_BP_all_spp_En[GO_APOC_BP_all_spp_En$Fisher<0.05,]
```

### Nvec - shared between all three spp 

##### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
nvec_gene2go<-read.delim("../data/annotations/Nvec_gene2go.tab", sep="\t")
```

Filter by shared orthogroups and species
```{r}
upset_targets_nvec <- upset_targets %>%
  filter(
    species == "Nvec",
    Nvec == 1,
    Ahya == 1,
    Apoc == 1)
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
nvec_clust_genes <- as.character(unique(upset_targets_nvec$target))

# All genes 
nvec_all_genes <- as.character(nvec_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
nvec_GeneList <- factor(as.integer(nvec_all_genes %in% nvec_clust_genes))
names(nvec_GeneList) <- nvec_all_genes
```

Read in gene-to-go-mappings
```{r}
nvec_gene2go_topgo<-readMappings("../data/annotations/Nvec_gene2go.tab", IDsep=";", sep="\t")
```

##### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_NVEC_BP_all_spp <-new("topGOdata", ontology="BP", gene2GO=nvec_gene2go_topgo, allGenes=nvec_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_NVEC_BP_all_spp_FE <- runTest(GO_NVEC_BP_all_spp, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_NVEC_BP_all_spp_En <- GenTable(GO_NVEC_BP_all_spp, Fisher = GO_NVEC_BP_all_spp_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_NVEC_BP_all_spp_En$Fisher<-as.numeric(GO_NVEC_BP_all_spp_En$Fisher)
GO_NVEC_BP_all_spp_En_sig<-GO_NVEC_BP_all_spp_En[GO_NVEC_BP_all_spp_En$Fisher<0.05,]
```

### Ahya - shared between all three spp 

##### Make list of genes for TopGO

Read in gene to go information (generated above)
```{r}
ahya_gene2go<-read.delim("../data/annotations/Ahya_gene2go.tab", sep="\t")
```

Filter by shared orthogroups and species
```{r}
upset_targets_ahya <- upset_targets %>%
  filter(
    species == "Ahya",
    Nvec == 1,
    Ahya == 1,
    Apoc == 1)

upset_targets_ahya$target <- gsub("-RA", "", upset_targets_ahya$target)
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
ahya_clust_genes <- as.character(unique(upset_targets_ahya$target))

# All genes 
ahya_all_genes <- as.character(ahya_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
ahya_GeneList <- factor(as.integer(ahya_all_genes %in% ahya_clust_genes))
names(ahya_GeneList) <- ahya_all_genes
```

Read in gene-to-go-mappings
```{r}
ahya_gene2go_topgo<-readMappings("../data/annotations/Ahya_gene2go.tab", IDsep=";", sep="\t")
```

##### Biological Processes 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_AHYA_BP_all_spp <-new("topGOdata", ontology="BP", gene2GO=ahya_gene2go_topgo, allGenes=ahya_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_AHYA_BP_all_spp_FE <- runTest(GO_AHYA_BP_all_spp, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_AHYA_BP_all_spp_En <- GenTable(GO_AHYA_BP_all_spp, Fisher = GO_AHYA_BP_all_spp_FE, orderBy = "Fisher", numChar = 51, topNodes = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_AHYA_BP_all_spp_En$Fisher<-as.numeric(GO_AHYA_BP_all_spp_En$Fisher)
GO_AHYA_BP_all_spp_En_sig<-GO_AHYA_BP_all_spp_En[GO_AHYA_BP_all_spp_En$Fisher<0.05,]
```

### Enriched terms for piRNA overlap for all 3 spp 

Find shared enriched terms 
```{r}
# Add species labels and keep only needed columns
ahya <- GO_AHYA_BP_all_spp_En_sig %>%
  mutate(species = "Ahya") %>%
  dplyr::select(species, GO.ID, Term, Fisher, Significant)

apoc <- GO_APOC_BP_all_spp_En_sig %>%
  mutate(species = "Apoc") %>%
  dplyr::select(species, GO.ID, Term, Fisher, Significant)

nvec <- GO_NVEC_BP_all_spp_En_sig %>%
  mutate(species = "Nvec") %>%
  dplyr::select(species, GO.ID, Term, Fisher, Significant)

# Shared GO IDs across all three
shared_ids <- purrr::reduce(
  list(ahya$GO.ID, apoc$GO.ID, nvec$GO.ID),
  intersect)
shared_ids
```

Filter by shared GO terms 
```{r}
# Bind all three, then filter to shared GO terms
go_all <- bind_rows(ahya, apoc, nvec)

go_shared <- go_all %>%
  filter(GO.ID %in% shared_ids)
```

Plot! 
```{r}
ggplot(go_shared, aes(x = species, y = Term, color = Fisher)) +
  geom_point(size = 10) +
  scale_fill_viridis_c(option = "D", name = expression('-log'[10]*'(Fisher p-value)')) +
  #facet_wrap(~ontology, scales = "free_y", ncol = 1) +
  xlab("Species") +
  ylab("GO Term") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
        strip.text = element_text(size = 12, face = "bold"))
```

