---
title: "tDRs"
author: "Jill Ashey"
date: "2026-09-08"
output: html_document
---

Plot tRNA-derived RNAs (or tRNA fragments?) across species 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(cowplot)
```

Read in data 
```{r}
files <- c("../output/tDR/ahya_tdr_counts.csv", "../output/tDR/apoc_tdr_counts.csv", "../output/tDR/nvec_tdr_counts.csv")
species <- c("Acropora hyacinthus", "Astrangia poculata", "Nematostella vectensis")

# Read, add species column, and combine
combined_tdr <- purrr::map2_dfr(files, species, ~ {
  read_csv(.x, show_col_types = FALSE) %>%
    mutate(Species = .y)
})
```

Summarize by species and type 
```{r}
species_tdr_summary <- combined_tdr %>%
  group_by(Species, Type) %>%
  summarise(Total_Count = sum(Count), .groups = "drop") %>%
  group_by(Species) %>%
  mutate(Percentage = (Total_Count / sum(Total_Count)) * 100)
```

Plot! 
```{r}
# Set levels
species_tdr_summary$Type <- factor(species_tdr_summary$Type, levels = c("5' tDR", "i-tDR", "3' tDR"))
species_tdr_summary$Species <- factor(species_tdr_summary$Species, levels = c("Acropora hyacinthus", "Astrangia poculata", "Nematostella vectensis"))

tdr_plot <- ggplot(species_tdr_summary, aes(x = Species, y = Percentage, fill = Type)) +
  geom_bar(stat = "identity", width = 0.8, color = "black", position = position_stack(reverse = T)) +
    scale_y_continuous(
        expand = c(0, 0),
        breaks = seq(0, 100, by = 10),
        limits = c(0, 105)) +

    scale_fill_manual(values = c(
    "5' tDR" = "darkblue",  
    "3' tDR" = "darkred",   
    "i-tDR"  = "darkolivegreen4")) +
  ylab("Percentage (%)") +
  xlab("") +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  theme_classic() + 
  theme(axis.text = element_text(color = "black", size = 20),
    axis.title = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(face = "italic"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 15),
    legend.key.size = unit(0.6, "cm")) +
  theme(
    aspect.ratio = 1.2, # Prevents vertical stretching
    legend.position = "bottom") +
  guides(fill = guide_legend(reverse = TRUE)); tdr_plot

ggsave("../output/tDR/tDR_percentage.pdf",
       tdr_plot, width = 8, height = 6, units = "in", dpi = 600, bg = "white")
ggsave("../output/tDR/tDR_percentage.png",
       tdr_plot, width = 8, height = 6, units = "in", dpi = 600, bg = "white")
```

## tRNA position

Set colors 
```{r}
trna_colors <- c(
  "GlyGCC" = "#0072B2",  # Deep Blue
  "AspGTC" = "#D55E00",  # Vermilion / Red-Orange
  "LysCTT" = "#CC79A7",  # Pink / Magenta
  "GluCTC" = "#E69F00",  # Amber Yellow
  "HisGTG" = "#009E73",  # Teal / Green
  "GluTTC" = "#F0E442",  # Bright Yellow
  "ThrGGT" = "#56B4E9",  # Sky Blue
  "ThrAGT" = "#785EF0",  # Bright Purple
  "AlaGGC" = "#882255",  # Wine / Deep Red
  "MetCAT" = "#117733"   # Dark Forest Green
)

theme_custom <- theme_classic(base_size = 9) +
  theme(
    text = element_text(family = "sans"),
    axis.text = element_text(color = "black", size = 18),
    axis.title = element_text(size = 21, face = "bold"),
    plot.title = element_text(face = "italic", size = 24, hjust = 0.5),
    #panel.border = element_rect(color = "grey60", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    #panel.grid.major.x = element_line(color = "grey90", linewidth = 0.3),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 15),
    legend.key.size = unit(0.4, "cm"),
    legend.position = c(0.75, 0.75))
    #plot.margin = margin(2,2,2,2))
```


### Positional coverage - Ahya

Read in df
```{r}
ahya_trna <- read_csv("../output/tDR/ahya_trna_positional_coverage.csv")
```

Extract tRNA and anticodon family 
```{r}
ahya_trna_clean <- ahya_trna %>%
  mutate(isoacceptor = str_extract(tRNA, "[A-Z][a-z]{2}[A-Z]{3}")) %>%
  filter(!is.na(isoacceptor)) %>%
  group_by(isoacceptor, Position) %>%
  summarise(
    Total_RPM = sum(RPM),
    Total_Raw = sum(Raw_Count),
    .groups = "drop") %>%
  group_by(isoacceptor) %>%
  filter(Total_Raw > 0) %>%
  ungroup()

length(unique(ahya_trna_clean$isoacceptor))
```

Identify most expressed tRNA isoacceptors 
```{r}
ahya_top_trnas <- ahya_trna_clean %>%
  group_by(isoacceptor) %>%
  summarise(max_rpm = max(Total_RPM)) %>%
  arrange(desc(max_rpm)) %>%
  slice_head(n = 4) %>%
  pull(isoacceptor)
```

Plot for Ahya
```{r}
# Filter for top tRNAs or specific ones (e.g., c("LeuTAG", "GlyGCC", "GluCTC"))
ahya_trna_plot <- ahya_trna_clean %>%
  filter(isoacceptor %in% ahya_top_trnas)

ahya_plot <- ggplot(ahya_trna_plot, aes(x = Position, y = Total_RPM, color = isoacceptor)) +
  geom_line(linewidth = 1.1, alpha = 0.85) +
  #facet_wrap(~ isoacceptor, scales = "free_y", ncol = 3) +
  scale_x_continuous(limits = c(1, 85), breaks = seq(0, 80, by = 20)) +  labs(
    x = "Mature tRNA Nucleotide Position",
    y = "Coverage Normalized \nto Total Reads (RPM)",
    color = "Isoacceptor",
    title = "Acropora hyacinthus"
  ) +
  scale_color_manual(values = trna_colors) +
  theme_classic() +
  theme_custom; ahya_plot
```

### Positional coverage - Apoc

Read in df
```{r}
apoc_trna <- read_csv("../output/tDR/apoc_trna_positional_coverage.csv")
```

Extract tRNA and anticodon family 
```{r}
apoc_trna_clean <- apoc_trna %>%
  mutate(isoacceptor = str_extract(tRNA, "[A-Z][a-z]{2}[A-Z]{3}")) %>%
  filter(!is.na(isoacceptor)) %>%
  group_by(isoacceptor, Position) %>%
  summarise(
    Total_RPM = sum(RPM),
    Total_Raw = sum(Raw_Count),
    .groups = "drop") %>%
  group_by(isoacceptor) %>%
  filter(Total_Raw > 0) %>%
  ungroup()

length(unique(apoc_trna_clean$isoacceptor))
```

Identify most expressed tRNA isoacceptors 
```{r}
apoc_top_trnas <- apoc_trna_clean %>%
  group_by(isoacceptor) %>%
  summarise(max_rpm = max(Total_RPM)) %>%
  arrange(desc(max_rpm)) %>%
  slice_head(n = 4) %>%
  pull(isoacceptor)
```

Plot for Apoc
```{r}
# Filter for top tRNAs or specific ones (e.g., c("LeuTAG", "GlyGCC", "GluCTC"))
apoc_trna_plot <- apoc_trna_clean %>%
  filter(isoacceptor %in% apoc_top_trnas)

apoc_plot <- ggplot(apoc_trna_plot, aes(x = Position, y = Total_RPM, color = isoacceptor)) +
  geom_line(linewidth = 1.1, alpha = 0.85) +
  #facet_wrap(~ isoacceptor, scales = "free_y", ncol = 3) +
  scale_x_continuous(limits = c(1, 85), breaks = seq(0, 80, by = 20)) +  labs(
    x = "Mature tRNA Nucleotide Position",
    y = "Coverage Normalized \nto Total Reads (RPM)",
    color = "Isoacceptor",
    title = "Astrangia poculata"
  ) +
  scale_color_manual(values = trna_colors) +
  theme_classic() +
  theme_custom; apoc_plot
```

### Positional coverage - Nvec

Read in df
```{r}
nvec_trna <- read_csv("../output/tDR/nvec_trna_positional_coverage.csv")
```

Extract tRNA and anticodon family 
```{r}
nvec_trna_clean <- nvec_trna %>%
  mutate(isoacceptor = str_extract(tRNA, "[A-Z][a-z]{2}[A-Z]{3}")) %>%
  filter(!is.na(isoacceptor)) %>%
  group_by(isoacceptor, Position) %>%
  summarise(
    Total_RPM = sum(RPM),
    Total_Raw = sum(Raw_Count),
    .groups = "drop") %>%
  group_by(isoacceptor) %>%
  filter(Total_Raw > 0) %>%
  ungroup()

length(unique(nvec_trna_clean$isoacceptor))
```

Identify most expressed tRNA isoacceptors 
```{r}
nvec_top_trnas <- nvec_trna_clean %>%
  group_by(isoacceptor) %>%
  summarise(max_rpm = max(Total_RPM)) %>%
  arrange(desc(max_rpm)) %>%
  slice_head(n = 4) %>%
  pull(isoacceptor)
```

Plot for Nvec
```{r}
# Filter for top tRNAs or specific ones (e.g., c("LeuTAG", "GlyGCC", "GluCTC"))
nvec_trna_plot <- nvec_trna_clean %>%
  filter(isoacceptor %in% nvec_top_trnas)

nvec_plot <- ggplot(nvec_trna_plot, aes(x = Position, y = Total_RPM, color = isoacceptor)) +
  geom_line(linewidth = 1.1, alpha = 0.85) +
  #facet_wrap(~ isoacceptor, scales = "free_y", ncol = 3) +
  scale_x_continuous(limits = c(1, 85), breaks = seq(0, 80, by = 20)) +  labs(
    x = "Mature tRNA Nucleotide Position",
    y = "Coverage Normalized \nto Total Reads (RPM)",
    color = "Isoacceptor",
    title = "Nematostella vectensis"
  ) +
  scale_color_manual(values = trna_colors) +
  theme_classic() +
  theme_custom; nvec_plot
```

## Plot all tDR things 

```{r}
# left col 
left <- plot_grid(
  tdr_plot, 
  ncol = 1, 
  align = "v",
  labels = c("A"), 
  label_size = 35)

# right col 
right <- plot_grid(
  ahya_plot,
  apoc_plot,
  nvec_plot,
  ncol = 1, 
  align = "v",
  labels = c("B", "C", "D"), 
  label_size = 35)

# combine 
full_figure <- plot_grid(
  left, 
  right, 
  ncol = 2, 
  rel_widths = c(0.45, 0.55)); full_figure

ggsave("../output/tDR/tDR_features.pdf",
       full_figure, width = 20, height = 16, units = "in", dpi = 600, bg = "white")
ggsave("../output/tDR/tDR_features.png",
       full_figure, width = 20, height = 16, units = "in", dpi = 600, bg = "white")
```




